% =============================================================================
% Abstract -- Refined framing per 25-prompt audit
% Drop-in include: \input{sections/abstract}
% =============================================================================
\begin{abstract}
We present TinkerRL-Bench, a systems-oriented empirical audit of 79 post-training
runs across tool-call formatting, GSM8K, MATH-style prompts, HumanEval-style
verification, multiple model families, and several training stacks.  Our
training runner is a group-relative policy-gradient variant inspired by GRPO,
not a faithful reproduction of the canonical objective of
\citet{shao2024deepseekmath}: it does not implement the PPO ratio and clip,
does not retain a frozen reference policy for KL regularization, uses a
single optimizer step per rollout batch, and does not apply a completion-only
token mask (a post-hoc P1-A diagnostic on Qwen3.5-4B measures 61.6\%--89.6\%
of the full-sequence loss magnitude to come from prompt tokens rather than
from the completion).  The contribution is diagnostic: characterizing the
structural boundaries of reward variance and stack identifiability that
govern whether critic-free RL produces a learning signal at all.

First, we show that GRPO hyperparameters are insufficient to specify training
dynamics.  Under nominally matched visible configurations (Qwen3-8B, group size
$G=8$, learning rate $10^{-5}$, GSM8K, 30 steps, seed 42), our Tinker runner
reaches 84.4\% last-10 training reward while TRL on Modal H100 reaches 5.0\%.
This 17$\times$ gap is not a framework ranking; it proves that backend,
sampler, reward parser, LoRA configuration, optimizer, and evaluator are
co-determinants of outcome in ways that hyperparameter tables do not capture.
Algorithm labels such as PPO, GRPO, and DPO are not meaningful experimental
treatments unless the full training stack is specified and controlled.

Second, we introduce Zero-Variance Fraction (ZVF) and Gradient Utilization (GU)
as actionable triage diagnostics for GRPO training signal.  ZVF measures the
fraction of prompts where all $G$ completions receive identical rewards,
producing zero gradient.  In our corpus, ZVF $\geq 80\%$ sustained over five
or more steps with reward $\leq 5\%$ correctly identifies all collapsed runs
(precision 1.0, recall 1.0 on a 22-run validation set).  Importantly, ZVF is
task-driven rather than model-scale-driven: tool-use experiments saturate at
ZVF = 100\% regardless of model size, while GSM8K experiments average ZVF =
8.5\%.  The zero-order ZVF--reward correlation ($r = -0.769$, $p = 0.0008$) is
largely tautological on binary-reward tasks (since ZVF$(p,G) = p^G + (1-p)^G$),
and the partial correlation controlling for current reward at lag 5 is only
$r = 0.059$ ($p = 0.096$).  We therefore treat ZVF/GU as regime-identifying
triage (cold-start collapse vs. saturation vs. healthy contrast), not as a
calibrated performance predictor.

Third, we show that training reward conflates with held-out capability in most
prior work.  The clean held-out GSM8K control (Qwen3-8B, paired evaluation on
a 200-example held-out slice not used during training) improves only from
82.0\% to 83.3\% ($p = 0.26$), a small and non-significant gain.  Held-out
evaluation of the ten checkpoints whose training \emph{last-10} reward is
highest yields an 87.2\%--95.0\% accuracy band, but this is a checkpoint-selection
analysis: selecting by training reward introduces over-optimism.  The clearest
example is four Llama-3.3-70B-Instruct seeds whose training \emph{last-10}
values range from 95.0\% to 98.1\% yet all land on 95.0\% held-out accuracy,
demonstrating that \emph{last-10} variance reflects Monte-Carlo sampling noise
rather than genuine generalization differences.

We do not claim GRPO universally improves reasoning.  We do not claim ZVF
predicts final performance.  We do not claim PPO is inferior to GRPO.  We do
not claim our benchmark establishes scaling laws (the exponential saturation
fits have mean $R^2 = 0.210$ on short traces).  Our contribution is strictly
diagnostic: the structural boundaries of reward variance and stack
identifiability that determine whether critic-free group-relative RL produces
a learning signal.
\end{abstract}