% =============================================================================
% Anonymous abstract -- three-claim framing
% =============================================================================
\begin{abstract}
We present a systems-oriented empirical audit of 79 post-training runs
across tool-call formatting, GSM8K, MATH-style prompts, HumanEval-style
verification, multiple model families, and several training stacks.  Our
training runner is a group-relative policy-gradient variant inspired by GRPO,
not a faithful reproduction of the canonical objective of
\citet{shao2024deepseekmath}: it does not implement the PPO ratio and clip,
does not retain a frozen reference policy for KL regularization, uses a
single optimizer step per rollout batch, and does not apply a completion-only
token mask (a post-hoc P1-A diagnostic on Qwen3.5-4B measures 61.6\%--89.6\%
of the full-sequence loss magnitude to come from prompt tokens rather than
from the completion).  The contribution is therefore not a
leaderboard claim about ``GRPO''; it is a diagnostic account of when a
critic-free group-relative RL loop has a usable learning signal and when
apparent training-reward gains fail to transport to held-out capability.

First, Zero-Variance Fraction (ZVF) and Gradient Utilization (GU) are best
treated as early triage diagnostics rather than calibrated performance
predictors.  High ZVF with low reward flags cold-start collapse; high ZVF
with high reward flags saturation; low ZVF indicates usable within-group
contrast.  A fixed first-five-step rule catches both collapsed runs in a
22-run de-duplicated validation set with no false positives, but only two
positive failures are observed, and continuous ZVF ranking is weak.
Second, the relevant mechanism is mixed-reward sampling probability:
\[
P(\text{usable}) \approx \frac{1}{N}\sum_x \left[1 - (1-p_x)^G - p_x^G\right].
\]
The familiar $1/G$ rule is only an expected-one-success heuristic.  Third,
PPO/GRPO/DPO labels are incomplete experimental treatments unless the model
family, backend, reward parser, sampler, optimizer, checkpoint rule, and
evaluation harness are specified.  Our Qwen PPO/GRPO comparison is
artifact-sensitive, and our Llama comparison favors PPO; neither supports a
universal algorithm ranking.

Several reward environments in this study measure proxies rather than
end-to-end capability: the HumanEval-style sandbox executes completions with
Python built-ins removed (rejecting legitimate uses of \texttt{len},
\texttt{range}, \texttt{sum}); the tool-use reward scores JSON well-formedness,
tool-name selection, and argument-key presence without executing tools; the
MATH reward combines \texttt{\textbackslash boxed\{\}} format credit with
final-answer matching; and the MATH-500 and HumanEval prompt pools used in the
training loop are loaded from the \texttt{test} split of their respective
datasets and therefore function as reward-environment probes rather than
generalization tests.  Held-out GSM8K, evaluated on a separate 200-example
slice that was not used during training, improves only from $82.0\%$ to
$83.3\%$ ($p=0.26$), so we do not claim a significant held-out math effect.
The defensible conclusion is that this runner behaves as a reward-contrast
amplifier: useful when the policy already samples distinguishable outcomes
under a verifier that rewards them, brittle when the reward distribution is
flat or when the verifier penalizes legitimate completions.
\end{abstract}
