% =============================================================================
% Introduction -- claim-to-evidence audit structure
% Drop-in include: \input{sections/intro}
% =============================================================================
\section{Introduction}
\label{sec:intro}

Reinforcement learning post-training now anchors many claims about language
model reasoning and alignment~\citep{ouyang2022training,guo2025deepseekr1nature}.
PPO~\citep{schulman2017proximal}, GRPO~\citep{shao2024deepseekmath}, and
DPO~\citep{rafailov2024direct} are often compared as if the algorithm label
were the full treatment.  In practice, the outcome also depends on the base
model, instruction tuning, reward parser, group construction, sampler,
backend, LoRA configuration, checkpoint selection, and evaluation harness.
This is the LLM post-training version of the reproducibility problem that
Henderson et al. identified in classical deep RL~\citep{henderson2018deep}.

\ours{} studies this problem as a diagnostic audit rather than as an
algorithm leaderboard.  The corpus contains 79 runs across structured
tool-call emission, GSM8K, MATH-style tasks, HumanEval-style binary
verification, small and larger model families, and several training stacks.
The central question is not whether GRPO broadly improves reasoning; it is
when a critic-free group-relative loop has a reward signal worth optimizing,
how that signal should be interpreted, and which implementation details must
be reported before algorithm comparisons are meaningful.

Before presenting results, we state three deliberate scoping choices.
(i) \textbf{Algorithm fidelity.}  The training runner is a group-relative
policy-gradient variant: it retains group-normalized advantages and the
critic-free structure of GRPO, but does not implement the PPO probability
ratio and clip~\citep{schulman2017proximal}, does not retain a frozen reference
policy for KL regularization, does one optimizer step per rollout batch, and
does not apply a completion-only token mask (a P1-A diagnostic on
Qwen3.5-4B measures 61.6\%--89.6\% of the full-sequence loss magnitude from
prompt tokens rather than completions).  Claims in this paper refer to this
runner, not to canonical GRPO.
(ii) \textbf{Evaluator scope.}  The HumanEval, tool-use, and MATH rewards
measure sandboxed execution with Python built-ins removed, JSON schema
compliance, and boxed-answer formatting respectively; they are proxies, not
end-to-end capability metrics.
(iii) \textbf{Probe vs. generalization.}  The MATH-500 and HumanEval prompt
pools in the training loop are loaded from the \texttt{test} splits and are
used as reward-environment probes, not as held-out generalization tests.  The
only genuinely held-out capability evaluation in this paper is a separate
GSM8K 200-example slice not used during training.

\begin{table}[t]
\centering
\caption{Claim--evidence--verdict hierarchy used throughout the paper.}
\label{tab:claim_evidence_verdict}
\small
\begin{tabular}{@{}p{0.27\linewidth}p{0.18\linewidth}p{0.33\linewidth}p{0.16\linewidth}@{}}
\toprule
\textbf{Claim} & \textbf{Status} & \textbf{Headline evidence} & \textbf{Verdict}\\
\midrule
C1: Reward contrast governs the group-relative learning signal. &
Main diagnostic contribution &
Mixed-group probability; early-run ZVF/GU; high-ZVF cold-start and saturation cases; measured base-hit-rate audits; 22-run first-five-step collapse triage. &
ZVF/GU are useful triage diagnostics, not calibrated performance predictors.\\
C2: Online reward is not held-out capability. &
Credibility anchor and central negative result &
Held-out GSM8K (Qwen3-8B, paired control): $82.0\% \to 83.3\%$, $p=0.26$; tool-use rewards are schema proxies; HumanEval/MATH harnesses are limited. &
Training reward supports dynamics claims; held-out evaluation is required for capability claims.\\
C3: Algorithm labels and visible GRPO hyperparameters are incomplete treatments. &
Methodological contribution &
Nominally matched config (Qwen3-8B, $G{=}8$, lr${=}10^{-5}$, GSM8K, 30 steps, seed 42): Tinker 84.4\% vs TRL 5.0\% last-10 reward; Qwen artifact sensitivity; Llama PPO-favored row. &
Report stack-conditioned claims, not universal framework or algorithm rankings.\\
\bottomrule
\end{tabular}
\end{table}

\subsection{Thesis}
This paper does not show that GRPO broadly improves reasoning.  The clean
held-out GSM8K control shows only a small, non-significant gain.  The thesis
is narrower and more useful: in a critic-free group-relative RL loop, the
existence of a learning signal is governed by reward diversity inside sampled
groups.  ZVF/GU expose two failure modes--cold-start collapse and saturation--
and help decide whether to stop, warm-start, densify the reward, increase group
size, or change task difficulty.  The broader experimental campaign is
interpreted through this evidence hierarchy: held-out evaluations support
capability claims; training rewards support dynamics claims; partial and proxy
runs support hypotheses only; and PPO/GRPO/DPO labels are not experimental
treatments unless the full post-training stack is specified.

\subsection{Claim 1: Reward contrast governs the learning signal.}
Zero-Variance Fraction (ZVF) measures the fraction of prompts where all $G$
completions receive identical rewards, producing zero group-relative advantage.
Gradient Utilization (GU) $=1-\mathrm{ZVF}$ measures the fraction of prompts
providing informative contrast.
High ZVF with low reward means cold-start collapse: all sampled answers are
wrong and the standard group-relative advantage is flat.
High ZVF with high reward means saturation: the task may already be solved
under the training reward, and further rollouts waste compute.
Low ZVF means the current policy still samples both successes and failures,
so the group-relative update has contrast to exploit.

Under binary rewards, if the current policy succeeds with success rate $p_x$
for prompt $x$ and group size is $G$, the probability of a usable mixed group is
\[
P(\text{usable}) \approx \frac{1}{N}\sum_x \left[1 - (1-p_x)^G - p_x^G\right].
\]
The familiar $p_x > 1/G$ rule is only an expected-one-success heuristic, not
a threshold theorem.  In a 22-run de-duplicated validation set, a fixed rule
using only the first five logged steps catches both collapsed runs with no
false positives.  Continuous ranking by early ZVF is weak, so the claim is
triage rather than forecasting.  Importantly, ZVF is task-driven, not
model-scale-driven: tool-use experiments saturate at ZVF = 100\% regardless of
model size, while GSM8K experiments average ZVF = 8.5\%.  The zero-order
ZVF--reward correlation ($r = -0.769$, $p = 0.0008$) is largely tautological
on binary-reward tasks (since ZVF$(p,G) = p^G + (1-p)^G$), and the partial
correlation controlling for current reward at lag 5 is only $r = 0.059$
($p = 0.096$).  ZVF/GU are regime-identifying diagnostics, not calibrated
performance predictors.

\subsection{Claim 2: Training reward is not capability.}
Online reward curves measure the reward environment that produced the update.
They should not be upgraded into general reasoning or tool-use claims without
matched held-out evaluation.  The clean GSM8K control (Qwen3-8B, paired 200-example
held-out slice) improves only from 82.0\% to 83.3\% ($p = 0.26$), so the paper
does not claim a significant held-out math effect.  Held-out evaluation of the
top-10 checkpoints (selected by training \emph{last-10} reward) yields an
87.2\%--95.0\% accuracy band, but this is a checkpoint-selection analysis:
selecting by training reward introduces over-optimism.  The clearest example
is four Llama-3.3-70B-Instruct seeds whose training \emph{last-10} values
range from 95.0\% to 98.1\% yet all land on 95.0\% held-out accuracy,
demonstrating that \emph{last-10} variance reflects Monte-Carlo sampling noise
rather than genuine generalization differences.  The positive tool-use rows
measure JSON validity, tool-name selection, and argument-key presence rather
than executed tool success; the HumanEval and MATH rows are bounded by their
sandbox and answer-parser choices.  These results remain useful as dynamics
evidence: they show when the rollout distribution supplies reward contrast
and when the verifier or parser becomes the bottleneck.

\subsection{Claim 3: Algorithm labels are under-specified treatments.}
The strongest safe version of the framework comparison is not that one backend
beats another.  It is that the phrase ``same GRPO config'' is under-specified.
Under nominally matched visible configurations (Qwen3-8B, group size $G=8$,
learning rate $10^{-5}$, GSM8K, 30 steps, seed 42), our Tinker runner reaches
84.4\% \emph{last-10} training reward while TRL on Modal H100 reaches 5.0\%
(peak 37.5\%).  This 17$\times$ gap does not prove that Tinker is better than
TRL; it shows that backend, sampler, reward parser, loss mask, KL/reference
handling, LoRA configuration, optimizer, checkpoint-selection rule, and
evaluator are part of the treatment.  Our Qwen PPO/GRPO comparison is
artifact-sensitive (the PPO last-10 value differs between the result ledger
and the statistical summary), and our Llama-3.1-8B-Instruct comparison favors
PPO over GRPO.  Algorithm-level claims require stack-conditioned reporting;
the treatment label is not enough.

\subsection{Contributions}
This paper contributes:
(i) a reward-diversity diagnostic for group-relative RL: ZVF and GU distinguish
cold-start collapse, saturation, and usable within-group contrast;
(ii) a mixed-group probability mechanism explaining why group size, base
success rate, warm starts, and reward density determine whether a learning
signal exists;
(iii) a negative capability result with evidence separation: online reward can
rise while the clean held-out GSM8K control remains non-significant
($82.0\% \to 83.3\%$, $p = 0.26$);
(iv) a stack-conditioned reporting protocol showing that PPO/GRPO/DPO labels
are insufficient without backend, reward parser, sampler, loss mask,
KL/reference handling, optimizer, LoRA, checkpoint, and evaluator details; and
(v) an auditable artifact ledger that separates online rewards, held-out
results, proxy tasks, partial runs, and artifact inconsistencies.
We release the code, run logs, and checkpoints at
\url{https://github.com/arvindcr4/tinker-rl-lab}.
