% =============================================================================
% Related Work v2 --- expanded related work for NeurIPS 2026 submission
% Replaces the Related Work section in paper/main.tex (sec:related).
% Grouped into five themes:
%   (1) RL algorithms for LLMs
%   (2) Reasoning RL (and process reward models)
%   (3) Tool-use RL
%   (4) Scale and reward-trajectory evidence for RL post-training
%   (5) Frameworks, evaluation harnesses, and reproducibility
% Every paragraph references entries defined in paper/references.bib.
% =============================================================================

\section{Related Work}
\label{sec:related}

\ours{} is not an optimizer paper in the style of DeepSeekMath, nor a
new zero-variance optimizer or prompt-selection rule in the style of RL-ZVP or
Online Difficulty Filtering.  It is a systems-oriented empirical audit of when
group-relative LLM post-training has a usable reward-contrast signal and when
its rewards become misleading evidence.  We therefore position the work along
five axes: policy-gradient \emph{algorithms} for language models, RL for
\emph{reasoning} (including process reward models and spurious-reward studies),
RL for \emph{tool use} and agentic behaviour, descriptive scale and
reward-trajectory studies for RL post-training, and the \emph{frameworks},
\emph{evaluation harnesses}, and reproducibility practices that instantiate
these algorithms.  The recurring
claim is that an algorithm label is not the full treatment: the model family,
reward parser, sampler, backend, mask, checkpoint rule, and evaluation split all
condition what a reported RL gain means.

% -----------------------------------------------------------------------------
\subsection{RL Algorithms for Large Language Models}
\label{sec:related:algos}

Proximal Policy Optimization (PPO)~\cite{schulman2017proximal} is the workhorse
algorithm behind InstructGPT-style RLHF~\cite{ouyang2022training,
christiano2017deeprlhf, stiennon2020summarize, bai2022hhrlhf, bai2022claudecai}
and has dominated production deployments for five years. Direct Preference
Optimization~\cite{rafailov2024direct} reframed preference learning as
contrastive classification and spawned a large family of RL-free alternatives,
including IPO~\cite{azar2024ipo}, KTO~\cite{ethayarajh2024kto},
SimPO~\cite{meng2024simpo}, ORPO~\cite{hong2024orpo}, and
SLiC~\cite{zhao2023slic}, all of which we consider as no-rollout baselines in
our cross-library protocol.

The current wave is defined by \emph{group-relative} policy gradient methods.
DeepSeekMath~\cite{shao2024deepseekmath} introduced GRPO together with a
7B math model trained from a math-heavy data pipeline; it framed GRPO as a
memory-efficient PPO variant that samples multiple completions per prompt,
uses group-relative rewards instead of a learned critic, and retains
PPO-style ratio/clip and reference-policy regularization.  DeepSeek-R1 later
popularised outcome-reward reasoning RL at much larger scale~\cite{guo2025deepseekr1,guo2025deepseekr1nature}.
This paper uses DeepSeekMath as the canonical anchor for what a faithful
``GRPO'' claim would require, but our runner is deliberately narrower: it
preserves the critic-free, group-relative reward-contrast mechanism while
omitting several canonical implementation choices, including probability-ratio
clipping, a frozen-reference KL term, and completion-only masking.  Our claims
are therefore mechanism-level and diagnostic rather than leaderboard-level: we
study when a group-relative loop has reward diversity to amplify, not whether a
particular DeepSeekMath recipe has been reproduced.  A second wave of refinements
dissects the biases of vanilla GRPO. Dr.~GRPO~\cite{liu2025drgrpo} removes a
response-length bias and a question-difficulty amplification bias by using a
constant loss normaliser and dropping the standard-deviation denominator in the
advantage. GDPO~\cite{liu2026gdpo} decouples group-relative normalisation per
reward component, preventing advantage collapse when rewards are multi-valued.
MAD~GRPO~\cite{wang2025madgrpo} replaces the standard-deviation advantage
scaler with the Median Absolute Deviation for heavy-tailed reward
distributions. G$^{2}$RPO~\cite{hu2026openvlthinker} forces per-prompt
advantages to match $\mathcal{N}(0,1)$ to fix reward-topology variance in
multimodal settings. Wu et al.~\cite{wu2025grpo_dpo} prove that GRPO is
``secretly DPO'' --- the group-level advantage is an implicit contrastive
objective --- and show that 2-GRPO retains $98.1\%$ of the performance of
16-GRPO at $12.5\%$ of the rollout budget. iGRPO~\cite{hatamizadeh2026igrpo}
adds a two-stage self-feedback loop that conditions subsequent rollouts on the
policy's best draft, reaching $85.6\%$ on AIME24. AERO~\cite{zhang2026aero}
combines adaptive rollout sizing, rejection sampling, and a Bayesian posterior
over prompt difficulty to eliminate zero-advantage dead zones, cutting RL
compute by $48\%$. Target Policy Optimization (TPO)~\cite{kaddour2026tpo}
decouples target-distribution construction from policy fitting: given scored
completions, TPO builds a target $q \propto p_{\text{old}} \exp(u)$ and fits
the policy via cross-entropy, giving a zero-gradient fixed point once the
policy matches the target. RGRA~\cite{carrino2026rgra} shows that PPO-style
clipping is unnecessary on top of group relative advantages, outperforming
GRPO on 17 of 27 comparisons. EFRame~\cite{wang2025eframe} augments GRPO with
exploration rollouts and experience replay for rare trajectories.

A parallel thread focuses on \emph{production-scale} refinements. GSPO, from
the Qwen team~\cite{qwen2025gspo}, redefines the importance ratio at the
sequence level (rather than token level) and was credited with stabilising the
Qwen3 series, especially its Mixture-of-Experts variants. DAPO, from
ByteDance~\cite{yu2025dapo}, contributes asymmetric Clip-Higher, dynamic
sampling, token-level loss normalisation, and overlong filtering, reaching
$50$ points on AIME~2024 with a 32B model. Kimi-K1.5~\cite{team2025kimi15,
team2025kimik2} and DeepSeek-V3~\cite{deepseek2024v3} both report their own
GRPO variants and online mirror-descent-style updates. Orthogonal
production variants include
Flow-GRPO~\cite{liu2025flowgrpo}, VAPO~\cite{bytedance2025vapo}, and
Dr.~SAC~\cite{ahmadian2024backtobasics}; the last argues that classical
REINFORCE with a value baseline, when carefully implemented, matches GRPO on
several RLHF tasks.

The closest algorithmic relatives are works that respond directly to
\emph{zero-variance} or zero-advantage groups.  RL-ZVP~\cite{rlzvp2025} argues
that all-correct and all-incorrect groups need not be discarded: it reshapes the
advantage with token-level signals so zero-variance prompts can still update the
policy.  Online Difficulty Filtering~\cite{bae2025online} takes the complementary
route of changing the prompt stream, selecting tasks where the current policy's
success probability is intermediate so sampled batches maintain high reward
variance.  These papers optimize around the same bottleneck that we measure.
The distinction is that \ours{} does not propose a replacement optimizer or
curriculum; it turns ZVF and Gradient Utilization (GU) into run-level diagnostics
for ordinary group-relative experiments.  High-ZVF/low-reward runs indicate
cold-start collapse; high-ZVF/high-reward runs indicate saturation; low ZVF
indicates usable within-group contrast.  Thus RL-ZVP and Online Difficulty
Filtering ask how to recover or select signal, while we ask whether a given
stack currently has signal and what evidential status its reward curve deserves.
Other proposals alter sampling or advantage estimation---AERO~\cite{zhang2026aero}
and GRESO~\cite{zhang2026greso} pre-screen prompts, CPPO~\cite{lin2025cppo}
prunes completions post-rollout, NGRPO~\cite{nan2025ngrpo} calibrates
all-incorrect groups, Scaf-GRPO~\cite{zhang2025scaffgrpo} injects hints, and
MO-GRPO~\cite{ichihara2025mogrpo} re-weights multi-objective rewards.  Theory
work such as Demystifying GRPO~\cite{zhou2026demystify}, MinPRO~\cite{lei2026minpro},
and SSPO~\cite{yang2025sspo} explains the estimator or granularity of the update;
our contribution is the complementary experimental accounting layer that decides
when those updates are being driven by real reward contrast.

% -----------------------------------------------------------------------------
\subsection{Reasoning RL and Process Reward Models}
\label{sec:related:reasoning}

The shift from outcome rewards to \emph{process} supervision has reshaped
reasoning RL. Let's-Verify-Step-by-Step~\cite{lightman2024stepbystep} showed
that step-level PRMs dramatically outperform outcome reward models on GSM8K
and MATH. Math-Shepherd~\cite{wang2024mathshepherd} derives process rewards
from auto-generated process annotations, avoiding costly human labels.
PRM800K~\cite{uesato2022prm800k} and PRMBench~\cite{song2024prmbench} provide
evaluation data for step-level scoring. OmegaPRM~\cite{luo2024omegaprm} uses
Monte-Carlo tree search to supervise process rewards without human-in-the-loop,
while ReasonEval~\cite{xia2024reasoneval} introduces reasoning-specific reward
models. Implicit PRM~\cite{yuan2024implicitprm} learns PRMs implicitly from
outcome labels, and VersaPRM~\cite{yuan2025versaprm} extends process rewards
to multi-domain tasks. These supervision signals are the backbone of recent
reasoning models, including OpenAI's o1/o3 family~\cite{openai2024o1,
openai2025o3}, DeepSeek-R1~\cite{guo2025deepseekr1, guo2025deepseekr1nature},
Qwen2.5-Math~\cite{yang2024qwen25math}, Qwen3~\cite{qwen2025qwen3},
Gemini~2.5 Pro's deliberative mode~\cite{google2025gemini25}, and
Kimi-K2~\cite{team2025kimik2}. Complementary results demonstrate that
self-consistency~\cite{wang2023selfconsistency}, tree-of-thought
search~\cite{yao2024tot}, and critic-style post-hoc
verifiers~\cite{mcaleese2024criticgpt} all benefit from or substitute for
process reward supervision. VerifyBench~\cite{zhang2025verifybench} and
A~Sober Look~\cite{hochlehnert2025sober} warn, however, that reward hacking
and evaluation artefacts can inflate perceived gains. Spurious
Rewards~\cite{shao2025spurious} is the closest negative control for our
interpretation: it shows that RLVR can yield large math-benchmark gains on
some model families even when the reward is random, format-only, incorrect, or
otherwise weakly aligned with correctness, and that these effects do not
necessarily transfer across model families.  Our proxy-reward results are read
under the same constraint.  Online reward curves are evidence about training
dynamics and reward-environment interaction; capability claims require held-out
or shifted evaluations.  This evidence hierarchy is therefore not a presentation
choice but the paper's response to the spurious-reward literature.

Reasoning-specific algorithmic contributions also continue to accumulate.
ReFT~\cite{luong2024reft} studies supervised fine-tuning versus RL for
reasoning. Let's Reward Step~\cite{zhang2025rewardstep} and
STaR~\cite{zelikman2022star} bootstrap reasoning chains from weak
supervisors. rStar-Math~\cite{guan2025rstarmath} combines MCTS-based PRMs
with self-evolved reasoning. ReST-EM~\cite{singh2024restem} alternates
expectation and maximisation steps over self-generated rationales.
Self-Taught Evaluator~\cite{wang2024stevaluator} adapts an LLM to score its
own reasoning chains. GRPO-Lead~\cite{lu2025grpolead}, LCPO~\cite{lu2025lcpo},
and LIMR~\cite{li2025limr} each tune reasoning depth with length-aware
rewards. Search-R1~\cite{jin2025searchr1} trains reasoning agents to
interleave search queries with chain-of-thought. TTRL~\cite{pu2025ttrl}
performs test-time RL against self-generated verifiers. We discuss the
Dr.~GRPO~\cite{liu2025drgrpo} and GSPO~\cite{qwen2025gspo} formulations as
related reference points for group-relative RL. Our own measurements across
0.6B--235B are primarily reward-harness diagnostics; held-out capability
claims are limited to the explicitly separated GSM8K evaluation.

% -----------------------------------------------------------------------------
\subsection{Tool-Use and Agentic RL}
\label{sec:related:tooluse}

Tool-use RL trains LLMs to interleave natural-language reasoning with external
API calls. Toolformer~\cite{schick2023toolformer} introduced self-supervised
tool-call insertion for a small set of APIs; Gorilla~\cite{patil2023gorilla}
retrieved from a large API catalogue; and ToolLLM~\cite{qin2024toolllm} scaled
training to thousands of real-world REST APIs. ReAct~\cite{yao2023react}
established the interleaved reason--act--observe pattern that most subsequent
agents inherit. RL specifically for tool use was pioneered by
ToolACE~\cite{liu2024toolace}, which couples a tool-calling reward with a
self-evolving synthesis pipeline; ToolRL~\cite{qian2025toolrl} applies GRPO
with verifiable tool-call rewards; ToRL~\cite{li2025torl} optimises the joint
reasoning--action trajectory. Further advances include
APIGen~\cite{lin2024apigen} (high-quality synthetic API data),
Nexus-Raven~\cite{nexusflow2024nexusraven}, and
Octopus~\cite{chen2024octopus}. Agentic RL extends tool use to long-horizon
tasks: WebGPT~\cite{nakano2021webgpt} and
WebShop~\cite{yao2022webshop} were the first large-scale web agents,
Reflexion~\cite{shinn2023reflexion} and Voyager~\cite{wang2023voyager} add
episodic self-reflection, and AutoGPT-style agents have been formalised by
AgentBench~\cite{liu2024agentbench}. 2025--2026 produced a wave of RL-trained
agents: SWE-RL~\cite{wei2025swerl} trains coding agents with execution
rewards; SWE-Gym~\cite{pan2025swegym} provides an RL training environment for
real GitHub issues; ARTIST~\cite{singh2025artist} trains tool-integrated
reasoning agents with GRPO; RAGEN~\cite{wang2025ragen} studies multi-turn
RL with tool access; OpenAI's Deep Research~\cite{openai2025deepresearch}
and Perplexity Deep Research~\cite{perplexity2025deepresearch} both report
GRPO-style training over web-browsing traces. Our audit corpus currently gives
its cleanest evidence on single-turn verifiable mathematical reasoning; tool-use
and browser-control probes are retained as auxiliary reward-environment
settings rather than end-to-end agentic capability claims.  We adopt Tinker's
\texttt{message\_with\_tool} interface as the neutral substrate for future
tool-use extensions and cite these works as concurrent settings in which
\ours{}'s ZVF measurements will need to be re-evaluated.

% -----------------------------------------------------------------------------
\subsection{Scale and Reward-Trajectory Evidence for RL Post-Training}
\label{sec:related:scaling}

Compute-optimal training of base language models is governed by
well-known scaling laws~\cite{kaplan2020scalinglaws, hoffmann2022chinchilla,
muennighoff2023datascaling}, but RL post-training is much less well
characterised. Hilton et al.~\cite{hilton2023rlscaling} first derived
compute-efficiency frontiers for RL in games and robotics. For RLHF,
Gao et al.~\cite{gao2023reward} characterise the over-optimisation curve of
reward models, and Rafailov et al.~\cite{rafailov2024rlhfscaling} provide
scaling laws for DPO and PPO that show a strong dependence on reward-model
quality. Nimmaturi et al.~\cite{nimmaturi2025scalinglaws} derive an
empirical scaling law specifically for GRPO, identifying three training
phases (slow start, rapid improvement, plateau) and showing that most
benefit is exhausted by roughly $80\%$ of one epoch. Liu et
al.~\cite{liu2025rlsubnetworks} find that RL fine-tuning updates only
$5$--$30\%$ of model parameters across seven algorithms and ten model
families, suggesting RL activates a sparse ``RL subnetwork.''
MiniCPM~\cite{hu2024minicpm} and the SLM survey of
Subramanian et al.~\cite{subramanian2025slm} characterise scaling in the
sub-8B regime that forms the bulk of our frontier. Schaeffer et
al.~\cite{schaeffer2024emergent} caution that apparent emergent abilities may
be metric artefacts, a concern that persists into RL. Our auxiliary
measurements are deliberately more modest: they provide descriptive
reward-trajectory probes from 0.6B to 235B across five families and four
frameworks.  We treat those traces as calibration context for future RL scaling
studies, not as a validated scaling law in this paper.

% -----------------------------------------------------------------------------
\subsection{Frameworks, Evaluation Harnesses, and Reproducibility}
\label{sec:related:frameworks}

The proliferation of algorithms has been matched by a proliferation of
training \emph{frameworks}: TRL~\cite{vonwerra2022trl},
OpenRLHF~\cite{hu2024openrlhf}, veRL~\cite{sheng2024verl},
NeMo-RL~\cite{nvidia2024nemorl}, DeepSpeed-Chat~\cite{yao2023deepspeedchat},
trlX~\cite{havrilla2023trlx}, Axolotl~\cite{axolotl2024}, and
\tinker{}~\cite{thinkingmachines2024tinker}. Each framework makes different
default choices about advantage normalisation, KL estimators, tokenisation
loss masking, rollout batching, and importance-sampling granularity. Those
choices are not incidental engineering details: they are part of the
realized treatment, so algorithm comparisons should be read as
stack-conditioned unless these factors are fixed or modeled.
On the inference side, vLLM~\cite{kwon2023vllm} and
SGLang~\cite{zheng2024sglang} provide the rollout backends on which all
modern frameworks depend; FlashAttention~\cite{dao2022flashattention} and
LoRA~\cite{hu2022lora} are load-bearing kernels.

Our methodological lineage is closest to the deep-RL
\emph{reproducibility} literature.  Henderson et al.~\cite{henderson2018deep}
showed that seeds, codebases, hyperparameters, and environment choices can
reverse apparent deep-RL conclusions; \texttt{rliable}~\cite{agarwal2021deep}
argued against relying on point estimates and introduced interval estimates,
interquartile means, and performance profiles; Patterson et
al.~\cite{patterson2024empirical} surveyed empirical design practices; and
Jordan et al.~\cite{jordan2024benchmarking} argued that full benchmark claims
are often too expensive to support with adequate power.  LLM RL post-training
adds additional treatment variables---tokenizer and chat template, reward
parser, verifier sandbox, rollout backend, KL/reference handling, loss mask,
LoRA target modules, checkpoint rule, and evaluation split---so the old
``implementation detail'' problem becomes a central scientific variable.
Pineau et al.~\cite{pineau2020improving} formalised NeurIPS reproducibility
criteria, and ACM's Artifact Review and Badging~\cite{acm2020badges}
established community standards for available, evaluated, and reproduced
artifacts.  Reproducibility studies in adjacent fields include
\cite{cavenaghi2023systematic, lynnerup2019survey, yamerenko2025generating,
paparella2023reproducibility, huang2022state, lopezibañez2021reproducibility,
leeser2026artifact, sedghpour2024artifact, colas2019hitchhiker}, alongside
Hoffman et al.'s Acme~\cite{hoffman2022acme}. Classical evaluations rest on
GSM8K~\cite{cobbe2021gsm8k}, MATH~\cite{hendrycks2021math},
AIME~\cite{maa2024aime}, MMLU~\cite{hendrycks2021mmlu},
HumanEval~\cite{chen2021codex}, MBPP~\cite{austin2021mbpp}, and
BBH~\cite{suzgun2023bbh}.  \ours{} ports these statistical traditions to the
LLM RL post-training regime: rather than offering a universal PPO/GRPO/DPO
ranking, it contributes stack-conditioned reporting, source-of-truth ledgers,
held-out-vs.-proxy claim separation, and ZVF/GU triage diagnostics within an
audit corpus spanning 0.6B to 235B parameters.

% =============================================================================
