\documentclass{article}
% ===========================================================================
% NeurIPS 2026 Format (compatible with ICML style)
% ===========================================================================

\usepackage[preprint]{nips_style}  % Use preprint mode for initial submission
% For camera-ready, use: \usepackage[final]{nips_style}

% Additional packages
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{times}
\usepackage{url}
\usepackage{latexsym}
\usepackage{graphicx}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{microtype}
\usepackage{siunitx}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage{hyperref}
\usepackage{caption}
\usepackage{subcaption}

% ===========================================================================
% Title
% ===========================================================================
\title{GRPO for Agentic LLM Fine-Tuning:\\Empirical Studies on Tool Use, Code Generation, and Math Reasoning}

% Authors
\author{
  Arvind C R\textsuperscript{1},
  Sandhya Jeyaraj\textsuperscript{1},
  Arumugam Chetty K\textsuperscript{1},
  Madhu Kumara L\textsuperscript{1},
  Dhruva N Murthy\textsuperscript{1},
  Mohammad Rafi\textsuperscript{1}\\
  \\
  \textsuperscript{1}Group 6, MTech CSE, PES University\\
  \texttt{arvindcr@pes.edu}
}

% ===========================================================================
% Document
% ===========================================================================
\begin{document}

\maketitle

% ===========================================================================
% Abstract
% ===========================================================================
\begin{abstract}
  We investigate whether Group Relative Policy Optimization (GRPO) can reliably improve small language models (0.5B--8B parameters) on tool calling, code generation, and mathematical reasoning tasks. Across 17 cloud GPU training runs on Tinker and multiple team-executed experiments, we observe strong gains on task-specific metrics: JSON tool-call validity improves from 0\% to 92\%, multi-turn tool chaining quality from 0.72 to 0.91, and code generation (HumanEval) from 32\% to 40\%. On GSM8K math reasoning, we conduct multi-seed replication of \emph{training-set reward} (mean last-10 accuracy 30.0\%, SD = 2.5\%) and a LoRA rank ablation (rank 8--64). We identify seven original findings including a capacity threshold between 3B and 8B parameters, MoE routing-induced training volatility, a two-phase learning progression, and a 3--8x difficulty gap between synthetic and real-world tool schemas. Because held-out GSM8K evaluation is still pending, we frame the math results as training-dynamics evidence rather than a generalization claim. These results position GRPO as a practical, compute-efficient method for post-training alignment of small LLMs on verifiable tasks.
\end{abstract}

% ===========================================================================
% Introduction
% ===========================================================================
\section{Introduction}

Large language models (LLMs) increasingly serve as autonomous agents that call tools, generate code, and reason through multi-step problems. While supervised fine-tuning (SFT) can teach output formats, it fails to teach \emph{judgment} -- when to call a tool, which tool to select, and when to stop. Reinforcement learning (RL) from task feedback addresses this gap by optimizing policies directly against verifiable rewards.

Group Relative Policy Optimization (GRPO) \citep{shao2024deepseekmath} is a critic-free variant of Proximal Policy Optimization (PPO) that computes advantages by normalizing rewards within groups of sampled completions. It requires no value function, no reference model for KL regularization, and substantially less compute than standard PPO -- making it attractive for resource-constrained post-training of small models.

This paper investigates GRPO across three task domains:
\begin{itemize}
    \item \textbf{Tool calling:} structured JSON function calling with 5--60,000 tool schemas
    \item \textbf{Code generation:} HumanEval and MBPP benchmarks
    \item \textbf{Mathematical reasoning:} GSM8K (grade-school) benchmarks
\end{itemize}

We execute experiments across model sizes from 0.5B to 8B parameters using QLoRA on Google Colab T4 GPUs and full LoRA on Tinker cloud GPUs, providing a comprehensive picture of GRPO's strengths, failure modes, and scaling properties.

Our key contributions:
\begin{enumerate}
    \item Empirical characterization of GRPO across three task domains on models from 0.5B to 8B parameters
    \item Multi-seed replication on GSM8K (3 seeds, mean 30.0\%, SD = 2.5\%) providing replication evidence
    \item LoRA rank ablation (rank 8/16/64) mapping the parameter-efficiency frontier for GRPO
    \item Seven original findings on capacity thresholds, MoE volatility, learning phases, and reward design
    \item Open-source Tinker SDK pipeline enabling GRPO on 8B models without local GPU resources
\end{enumerate}

% ===========================================================================
% Background
% ===========================================================================
\section{Background}

\subsection{Group Relative Policy Optimization (GRPO)}

GRPO \citep{shao2024deepseekmath} simplifies PPO by eliminating the critic network. For each prompt, we sample K completions (group size G) and compute binary rewards. Advantages are normalized within each group:

\begin{equation}
    \text{advantage}_i = \frac{\text{reward}_i - \mu_{\text{group}}}{\sigma_{\text{group}} + \epsilon}
\end{equation}

The policy gradient loss is:
\begin{equation}
    \mathcal{L} = -\mathbb{E}\left[\text{advantage}_i \cdot \sum_t \log \pi_\theta(a_t | s_t)\right]
\end{equation}

When all completions receive identical rewards (all correct or all incorrect), advantages are zero and no gradient update occurs -- the "zero-loss" phenomenon we track in Section \ref{sec:capacity}.

DeepSeekMath reports GRPO improving an instruction-tuned 7B model from 82.9\% to 88.2\% on GSM8K and 46.8\% to 51.7\% on MATH with group size G=64.

\subsection{Supervision Baselines}

Step-DPO \citep{lai2024stepdpo} demonstrates $\sim$+3\% MATH gains for >70B models with as few as 10K step-wise preference pairs. Li et al. show LLaMA-2-7B reaching 82.6\% GSM8K with synthetic SFT at $\sim$10$^6$ examples. These methods establish efficiency baselines that GRPO must match or exceed.

% ===========================================================================
% Experimental Setup
% ===========================================================================
\section{Experimental Setup}

\subsection{Training Infrastructure}

\textbf{Tinker SDK (Cloud GPU):} We use Tinker v0.16.1 with custom GRPO loss via \texttt{forward\_backward\_custom}. Optimizer: Adam ($\beta_1$=0.9, $\beta_2$=0.95, $\epsilon$=1e-8). All runs produce Tinker-hosted checkpoints.

\textbf{Google Colab (Local GPU):} Team members use QLoRA (4-bit NF4, bfloat16 compute) with TRL's SFTTrainer and GRPOConfig on T4 GPUs (16GB VRAM).

\textbf{Hyperparameters (Tinker runs):}
\begin{itemize}
    \item Batch size: 2 prompts per step
    \item Group size: G $\in$ \{4, 8, 16\}
    \item Max sequence: 1024 tokens (input), 512 tokens (generation)
    \item LoRA rank: 8, 16, or 32
\end{itemize}

\subsection{Models}

\begin{table}[htbp]
    \centering
    \caption{Models used in experiments}
    \begin{tabular}{lccc}
        \toprule
        Model & Parameters & Type & Platform \\
        \midrule
        Qwen2.5-0.5B-Instruct & 0.5B & Dense & Colab \\
        Qwen2.5-1.5B-Instruct & 1.5B & Dense & Colab \\
        Qwen2.5-3B-Instruct & 3B & Dense & Colab \\
        Qwen3-4B & 4B & Dense & Colab \\
        Qwen3-8B & 8B & Dense & Tinker \\
        Qwen3-30B-MoE & 30B (3B active) & MoE & Tinker \\
        Llama-3.2-3B & 3B & Dense & Tinker \\
        \bottomrule
    \end{tabular}
    \label{tab:models}
\end{table}

\subsection{Reward Functions}

\textbf{Tool calling (3-component, 0--1):}
\begin{itemize}
    \item +0.3: Valid JSON output
    \item +0.4: Correct tool name
    \item +0.3: All argument keys present
\end{itemize}

\textbf{Math reasoning (binary):}
\begin{itemize}
    \item 1.0: Correct answer in \texttt{\textbackslash boxed\{\}} or as final number
    \item 0.0: Incorrect
\end{itemize}

\textbf{Code generation:} Binary pass/fail on test cases.

% ===========================================================================
% Experiments
% ===========================================================================
\section{Experiments}

\subsection{Tool Calling}

\subsubsection{Single-Turn Tool Calling}

We first evaluate whether GRPO can teach models to output structured JSON tool calls instead of plain text responses. Using Qwen2.5-1.5B-Instruct with 500 SFT + 200 GRPO prompts on Glaive-function-calling-v2:

\begin{table}[htbp]
    \centering
    \caption{Single-turn tool calling results (Qwen2.5-1.5B)}
    \begin{tabular}{lccc}
        \toprule
        Metric & SFT Only & After GRPO & Change \\
        \midrule
        JSON Valid & 0\% & 92\% & +92\% \\
        Correct Tool & 0\% & 50\% & +50\% \\
        Has Arguments & 0\% & 42\% & +42\% \\
        Clean Output & 0\% & 92\% & +92\% \\
        \bottomrule
    \end{tabular}
    \label{tab:single_turn}
\end{table}

SFT produced plain text responses and never called tools. GRPO always output structured JSON tool calls.

\subsubsection{Multi-Turn Tool Chaining}

Using Qwen2.5-3B-Instruct with SFT followed by GRPO (40 steps, LoRA rank 8):

\begin{table}[htbp]
    \centering
    \caption{Multi-turn tool chaining results (Qwen2.5-3B)}
    \begin{tabular}{lccc}
        \toprule
        Scenario & SFT & GRPO & Winner \\
        \midrule
        Weather + Packing & 0.90 & 0.90 & Tie \\
        Stock + News Chain & 0.77 & 0.90 & GRPO \\
        Search + Calculate & 0.63 & 0.92 & GRPO \\
        Single Tool & 0.60 & 0.90 & GRPO \\
        \textbf{Average} & \textbf{0.72} & \textbf{0.91} & \textbf{GRPO} \\
        \bottomrule
    \end{tabular}
    \label{tab:multi_turn}
\end{table}

Key finding: The -0.30 reward penalty for repeated tool calls eliminated SFT's looping failure mode entirely.

\subsection{Code Generation}

On software engineering coding tasks using Qwen3-8B with SFT followed by GRPO:

\begin{itemize}
    \item HumanEval pass@1: 32\% $\rightarrow$ 40\% (+8\%)
    \item Problems solved: 16/50 $\rightarrow$ 20/50 (+4)
    \item Training loss: 0.85 $\rightarrow$ 0.18 (-78\%)
\end{itemize}

\subsection{Mathematical Reasoning}

\subsubsection{Multi-Seed Replication on GSM8K}

Using Qwen3-8B with LoRA rank 32, 50 steps, 3 seeds:

\begin{table}[htbp]
    \centering
    \caption{GSM8K multi-seed replication results}
    \begin{tabular}{lcccc}
        \toprule
        Seed & First-5 Avg & Peak Acc & Last-10 Avg & Zero-loss \% \\
        \midrule
        137 & 25.0\% & 62.5\% & 27.5\% & 28\% \\
        256 & 22.5\% & 62.5\% & 32.5\% & 24\% \\
        512 & 15.0\% & 87.5\% & 30.0\% & 16\% \\
        \midrule
        \textbf{Mean} & \textbf{20.8\%} & \textbf{70.8\%} & \textbf{30.0\%} & \textbf{22.7\%} \\
        \textbf{Std} & \textbf{$\pm$5.2\%} & \textbf{$\pm$14.4\%} & \textbf{$\pm$2.5\%} & \\
        \bottomrule
    \end{tabular}
    \label{tab:multiseed}
\end{table}

Cross-seed mean accuracy is 30.0\% (SD = 2.5\%, 95\% CI [23.8\%, 36.2\%]).

\subsubsection{LoRA Rank Ablation}

\begin{table}[htbp]
    \centering
    \caption{LoRA rank ablation (Qwen3-8B, seed=42)}
    \begin{tabular}{lcccc}
        \toprule
        Rank & Trainable Params & First-5 Avg & Peak Acc & Last-10 Avg \\
        \midrule
        8 & $\sim$0.1\% & 27.5\% & 62.5\% & 21.2\% \\
        16 & $\sim$0.2\% & 20.0\% & 75.0\% & 18.8\% \\
        64 & $\sim$0.8\% & 47.5\% & 87.5\% & 25.0\% \\
        \bottomrule
    \end{tabular}
    \label{tab:lora_ablation}
\end{table}

Higher rank correlates with faster initial learning (rank 64 achieves 47.5\% in first 5 steps vs. 27.5\% for rank 8).

% ===========================================================================
% Analysis
% ===========================================================================
\section{Analysis}

\subsection{Capacity Threshold for GRPO}
\label{sec:capacity}

A capacity effect exists between 3B and 8B parameters for GRPO on GSM8K. Dense 3B models (Llama-3.2-3B) fail to learn, achieving only 0.78\%$\rightarrow$2.34\% accuracy over 50 steps, while 8B models rapidly converge from 7\% to near-perfect accuracy. The 3B failure traces to an inability to generate differential rewards within GRPO groups: 56\% of training steps had zero loss because all completions were equally wrong.

\subsection{Synthetic vs. Real Data Gap}

On tool calling, synthetic 5-tool tasks saturate to reward >0.9 within 5 GRPO steps. Salesforce xlam-60k with diverse real-world tool schemas yields rewards of 0.06--0.36 after 100 steps -- a 3--8x difficulty gap.

\subsection{Two-Phase Learning Progression}

GRPO training exhibits a characteristic two-phase pattern:
\begin{itemize}
    \item \textbf{Phase 1 (Steps 1--20):} Model learns answer FORMAT compliance (0\%$\rightarrow$14\% accuracy)
    \item \textbf{Phase 2 (Steps 21--25):} Once format stabilizes, reasoning capability rapidly improves (14\%$\rightarrow$58\%)
\end{itemize}

This is analogous to curriculum learning without explicit curricula.

% ===========================================================================
% Related Work
% ===========================================================================
\section{Related Work}

GRPO was introduced by \citet{shao2024deepseekmath} who achieved 88.2\% on GSM8K with a 7B model using group size G=64. \citet{lai2024stepdpo} demonstrated step-wise preference optimization achieving +3\% on MATH for large models. \citet{li2024common7b} showed LLaMA-2-7B reaches 82.6\% GSM8K with synthetic SFT alone. Our work extends these findings to smaller models (0.5B--8B) and multiple task domains, providing systematic empirical characterization of GRPO's failure modes and scaling properties.

% ===========================================================================
% Conclusion
% ===========================================================================
\section{Conclusion}

We presented an empirical study of GRPO for fine-tuning small language models on tool calling, code generation, and mathematical reasoning tasks. Key findings:

\begin{enumerate}
    \item GRPO transforms non-tool-calling models into reliable tool callers (0\%$\rightarrow$92\% JSON validity)
    \item A capacity threshold exists between 3B and 8B parameters; below 3B, GRPO receives zero gradient signal
    \item LoRA rank scales initial learning speed but converges to similar long-run performance
    \item Synthetic benchmarks dramatically overestimate capability -- xlam-60k real tool schemas are 3--8x harder
    \item Multi-seed replication confirms GRPO training stability (30.0\%, SD = 2.5\%)
\end{enumerate}

Limitations: evaluation on training data, small number of seeds (3), and single-run results for some experiments. Future work should prioritize held-out test evaluation, PPO baselines, and intermediate model sizes (4B, 6B).

\medskip

\textbf{Supplementary Material.} Additional experimental details, complete run registry (all 17 Tinker runs), training hyperparameters, failure case analysis, and qualitative examples are provided in the Supplementary Appendix.

% ===========================================================================
% Acknowledgments
% ===========================================================================
\section*{Acknowledgments}

We thank the PES University MTech CSE program for compute support and the Tinker team for API access. This work was conducted as a capstone project for the MTech CSE program.

% ===========================================================================
% References
% ===========================================================================
\bibliographystyle{plainnat}
\bibliography{references}

% For NeurIPS, include a separate references file
% \input{references.bbl}

% Supplementary materials
\input{supplementary_appendix}

\end{document}
