% ===========================================================================
% Supplementary Appendix
% GRPO for Agentic LLM Fine-Tuning
% ===========================================================================

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{times}
\usepackage{url}
\usepackage{latexsym}
\usepackage{graphicx}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{microtype}
\usepackage{hyperref}
\usepackage{caption}

% ===========================================================================
% Supplementary Material
% ===========================================================================
\begin{document}

% ===========================================================================
\section{Complete Training Run Registry}
% ===========================================================================

\subsection{Tool-Use Experiments (Tinker Cloud GPU, Qwen3-8B)}

\begin{table}[htbp]
    \centering
    \caption{Complete tool-use GRPO runs on Tinker (Qwen3-8B, LoRA rank 32). All runs use Adam optimizer ($\beta_1$=0.9, $\beta_2$=0.95, $\epsilon$=1e-8), max sequence 1024 input / 512 generation.}
    \begin{tabular}{llcccccl}
        \toprule
        Run ID & Config & Group & Temp & Steps & Last-10 Reward & Status \\
        \midrule
        88ed2271 & Baseline: LR=3e-5 & 8 & 0.8 & 30 & 0.875 & Done \\
        2d488a85 & High LR: LR=1e-4 & 16 & 0.8 & 30 & 0.999 & Done \\
        5569c1fa & Low Temp: LR=3e-5 & 4 & 0.4 & 30 & 0.977 & Done \\
        22e9e5fd & xlam-60k: LR=3e-5 & 8 & 0.8 & 30 & 0.363 & Done \\
        386273f4 & Synthetic 5-tool & 4 & 0.8 & 100 & 0.825 & Done \\
        f63b618c & xlam-60k extended & 4 & 0.8 & 100 & 0.113 & Done \\
        1b01abef & MATH problems & 4 & 0.8 & 100 & 0.264 & Done \\
        \bottomrule
    \end{tabular}
    \label{tab:tool_runs}
\end{table}

\subsection{GSM8K Mathematical Reasoning Experiments (Tinker, Qwen3-8B)}

\begin{table}[htbp]
    \centering
    \caption{Complete GSM8K GRPO runs on Tinker. Multi-seed replication (Qwen3-8B), LoRA rank ablation, and 4B scaling experiment. All metrics are training-set reward.}
    \begin{tabular}{lllcccccc}
        \toprule
        Run ID & Model & Seed & Rank & LR & Steps & Peak & Last-10 \\
        \midrule
        \multicolumn{8}{l}{\emph{Multi-seed replication (Qwen3-8B, rank 32)}} \\
        899d909e & Qwen3-8B & 42 & 32 & 3e-5 & 50 & 62.5\% & 27.5\% \\
        5db4e965 & Qwen3-8B & 137 & 32 & 3e-5 & 50 & 62.5\% & 27.5\% \\
        aabb48cb & Qwen3-8B & 256 & 32 & 3e-5 & 50 & 62.5\% & 32.5\% \\
        99971b26 & Qwen3-8B & 512 & 32 & 3e-5 & 50 & 87.5\% & 30.0\% \\
        b3ba8df6 & Qwen3-8B & 999 & 32 & 3e-5 & 50 & 87.5\% & 35.0\% \\
        \midrule
        \multicolumn{8}{l}{\emph{LoRA rank ablation (Qwen3-8B, seed 42)}} \\
        ba2a1694 & Qwen3-8B & 42 & 8 & 3e-5 & 50 & 62.5\% & 21.2\% \\
        92ebcc48 & Qwen3-8B & 42 & 16 & 3e-5 & 50 & 75.0\% & 18.8\% \\
        9219771c & Qwen3-8B & 42 & 64 & 3e-5 & 50 & 87.5\% & 25.0\% \\
        \midrule
        \multicolumn{8}{l}{\emph{Extended run and scaling}} \\
        1cc20cec & Qwen3-8B & 42 & 32 & 5e-6 & 100 & 75.0\% & 27.5\% \\
        566747c0 & Qwen3.5-4B & 137 & 32 & 3e-5 & 50 & 100\% & 82.5\% \\
        \bottomrule
    \end{tabular}
    \label{tab:gsm8k_runs}
\end{table}

\subsection{Google Colab QLoRA Experiments}

\begin{table}[htbp]
    \centering
    \caption{Local QLoRA experiments on Google Colab (T4 GPU, 16GB VRAM).}
    \begin{tabular}{lllccl}
        \toprule
        Model & Dataset & Method & LoRA Rank & Steps & Result \\
        \midrule
        Qwen2.5-0.5B & 14 synthetic & SFT only & 16 & 10 & JSON 30\% \\
        Qwen2.5-1.5B & Glaive-v2 & SFT+GRPO & 16 & 700 & JSON 92\% \\
        Qwen2.5-3B & ToolBench-v1 & SFT+GRPO & 8 & 40 & 0.91 multi-turn \\
        Llama-3.2-3B & GSM8K & GRPO only & 32 & 50 & 2.34\% acc \\
        Qwen3-4B & GSM8K & SFT+GRPO & 32 & 50 & In progress \\
        \bottomrule
    \end{tabular}
    \label{tab:colab_runs}
\end{table}

\subsection{Advanced RL Algorithm Variants (Colab Notebook)}

We provide a reproducible Colab notebook (\texttt{advanced\_rl\_colab.ipynb}) implementing three non-standard RL algorithms for comparison with standard GRPO:

\begin{table}[htbp]
    \centering
    \caption{Advanced RL algorithm configurations (Qwen3-8B + LoRA rank 32 on GSM8K).}
    \begin{tabular}{llcccc}
        \toprule
        Algorithm & Key Change & $\beta$ (KL) & LR & Temp & Steps \\
        \midrule
        Standard GRPO & Baseline & 0.04 & 4e-5 & 1.0 & 50 \\
        Dr.\ GRPO & No KL + entropy bonus & 0.0 & 5e-5 & 1.0 & 50 \\
        DAPO & Overlong filter + high temp & 0.0 & 4e-5 & 1.2 & 50 \\
        DPO & Offline preference pairs & 0.1 & 5e-6 & -- & 3 epochs \\
        \bottomrule
    \end{tabular}
    \label{tab:advanced_rl}
\end{table}

\noindent\textbf{Dr.\ GRPO} \citep{deepseekr1} removes the KL divergence penalty and adds a token-level entropy bonus ($\lambda=0.01$) to prevent the entropy collapse documented in Section~5.3 of the main paper. \textbf{DAPO} \citep{yu2025dapo} further decouples alignment by using asymmetric advantage clipping and higher sampling temperature for diverse rollouts. \textbf{DPO} \citep{rafailov2024dpo} uses rejection-sampled preference pairs (correct vs.\ incorrect completions from the base model) for offline preference learning without on-policy rollouts.

% ===========================================================================
\section{Training Hyperparameters}
% ===========================================================================

\subsection{Tinker Cloud GPU Configuration}

\begin{table}[htbp]
    \centering
    \caption{Tinker SDK hyperparameters for all GRPO runs.}
    \begin{tabular}{ll}
        \toprule
        Parameter & Value \\
        \midrule
        SDK Version & v0.16.1 \\
        Optimizer & Adam \\
        $\beta_1$ & 0.9 \\
        $\beta_2$ & 0.95 \\
        $\epsilon$ & 1e-8 \\
        Weight Decay & 0 \\
        Batch Size & 2 prompts/step \\
        Group Size & 4, 8, or 16 \\
        Max Input Length & 1024 tokens \\
        Max Generation Length & 512 tokens \\
        Warmup & None (constant LR) \\
        Checkpoint Frequency & Final only \\
        \bottomrule
    \end{tabular}
    \label{tab:tinker_hyperparams}
\end{table}

\subsection{LoRA Configuration}

\begin{table}[htbp]
    \centering
    \caption{LoRA adapter targets and ranks used across experiments.}
    \begin{tabular}{ll}
        \toprule
        Parameter & Value \\
        \midrule
        Target Modules & q\_proj, k\_proj, v\_proj, o\_proj, \\
                        & gate\_proj, up\_proj, down\_proj \\
        LoRA Ranks Tested & 8, 16, 32, 64 \\
        Default Rank & 32 \\
        Alpha & 2 $\times$ rank (default) \\
        Dropout & 0.0 \\
        QLoRA (Colab) & 4-bit NF4, bfloat16 compute \\
        \bottomrule
    \end{tabular}
    \label{tab:lora_config}
\end{table}

% ===========================================================================
\section{Failure Case Analysis}
% ===========================================================================

\subsection{Three-Billion Parameter Failure Mode}

The Llama-3.2-3B model on GSM8K (run not shown in main paper) achieved only 2.34\% accuracy after 50 GRPO steps, barely above the 1.56\% random baseline. Analysis of training logs reveals the root cause: \textbf{56\% of training steps had zero gradient} because all G=4 sampled completions received identical rewards (all correct or all incorrect).

This ``zero-loss'' phenomenon indicates the 3B model lacks the reasoning capacity to generate diverse completion quality within a group. When all completions are equally wrong, advantages are zero and no learning occurs. The 8B model (Qwen3-8B) shows only 16--28\% zero-loss rate, enabling effective gradient updates.

\begin{table}[htbp]
    \centering
    \caption{Zero-loss step percentage across model sizes (lower is better).}
    \begin{tabular}{lccc}
        \toprule
        Model & Parameters & Zero-Loss \% & Final Acc \\
        \midrule
        Llama-3.2-3B & 3B & 56\% & 2.34\% \\
        Qwen3-8B (seed 512) & 8B & 16\% & 30.0\% \\
        Qwen3-8B (seed 137) & 8B & 28\% & 27.5\% \\
        Qwen3-8B (seed 256) & 8B & 24\% & 32.5\% \\
        \bottomrule
    \end{tabular}
    \label{tab:zero_loss}
\end{table}

\subsection{Comprehensive Zero-Loss Dynamics Across All GSM8K Runs}

Table~\ref{tab:zero_loss_comprehensive} reports zero-loss and zero-reward step frequencies across all available GSM8K training runs, providing the quantitative basis for the capacity-threshold and exploration-sparsity claims in the main paper.

\begin{table}[htbp]
    \centering
    \caption{Zero-loss and zero-reward dynamics across all GSM8K GRPO runs. Zero-loss indicates all-same-reward groups (no gradient signal). Zero-reward indicates all-incorrect completions. 3B data from Colab; 4B and 8B from Tinker.}
    \begin{tabular}{llccccc}
        \toprule
        Model & Config & Steps & Zero-Loss \% & Zero-Reward \% & Peak Acc & Last-10 Acc \\
        \midrule
        Llama-3.2-3B & seed=42, G=4 & 50 & 56\% & 56\% & 3.1\% & 2.3\% \\
        \midrule
        Qwen3.5-4B & seed=137, r=32 & 50 & 68\% & 0\% & 100\% & 82.5\% \\
        \midrule
        Qwen3-8B & seed=137, r=32 & 50 & 28\% & 24\% & 62.5\% & 27.5\% \\
        Qwen3-8B & seed=256, r=32 & 50 & 24\% & 24\% & 62.5\% & 32.5\% \\
        Qwen3-8B & seed=512, r=32 & 50 & 16\% & 10\% & 87.5\% & 30.0\% \\
        Qwen3-8B & seed=42, r=32 & 50 & 18\% & 14\% & 62.5\% & 27.5\% \\
        Qwen3-8B & seed=999, r=32 & 50 & 18\% & 16\% & 87.5\% & 35.0\% \\
        Qwen3-8B & seed=42, r=8 & 50 & 20\% & 14\% & 62.5\% & 21.2\% \\
        Qwen3-8B & seed=42, r=16 & 50 & 20\% & 18\% & 75.0\% & 18.8\% \\
        Qwen3-8B & seed=42, r=64 & 50 & 18\% & 16\% & 87.5\% & 25.0\% \\
        Qwen3-8B & lr=5e-6, 100 steps & 100 & 17\% & 15\% & 75.0\% & 27.5\% \\
        \bottomrule
    \end{tabular}
    \label{tab:zero_loss_comprehensive}
\end{table}

\textbf{Key patterns:}
\begin{itemize}
    \item The 3B model's 56\% zero-loss rate stems from zero-reward (all-incorrect) groups, confirming reward sparsity as the proximate cause.
    \item The 4B model's 68\% zero-loss rate stems from \emph{all-correct} groups (0\% zero-reward), indicating the opposite regime: the model solves problems so easily that group variance vanishes. This is productive zero-loss (learning saturated, not stalled).
    \item 8B models show 16--28\% zero-loss, with seed 512 showing the lowest rate and highest peak accuracy, suggesting exploration varies meaningfully across seeds.
    \item LoRA rank does not substantially affect zero-loss rates (18--20\% for r=8/16/64), supporting the paper's finding that rank affects learning speed but not the fundamental capacity to generate reward signal.
\end{itemize}

\subsection{Qwen3-30B-MoE Volatility Analysis}

The Qwen3-30B-MoE model reached 99\% peak GSM8K training-step accuracy in our internal runs but exhibited 2.43$\times$ higher step-to-step reward variance than the dense 8B model. Statistical testing (Levene's test) confirms this difference is significant: $F(1, 98) = 22.4$, $p = 7.0 \times 10^{-6}$.

This volatility stems from MoE routing: different experts activate for different tokens, creating non-smooth loss surfaces. The sparse activation pattern means small changes in the router can dramatically shift which experts process each token, amplifying training instability.

\begin{table}[htbp]
    \centering
    \caption{MoE vs. Dense training stability comparison.}
    \begin{tabular}{lccc}
        \toprule
        Model & Step Var (SD) & Peak Acc & Final Acc \\
        \midrule
        Qwen3-8B Dense & 0.12 & 87.5\% & 30.0\% \\
        Qwen3-30B-MoE & 0.29 & 99.0\% & 95.0\% \\
        Ratio (MoE/Dense) & 2.43$\times$ & --- & --- \\
        \bottomrule
    \end{tabular}
    \label{tab:moe_volatility}
\end{table}

\subsection{Policy Entropy, KL, and Group-Composition Diagnostics}
\label{sec:diagnostics_table}

The arithmetic task provides the richest telemetry logged during GRPO training. Table~\ref{tab:diagnostics} summarizes key diagnostic metrics at milestone training steps. Policy entropy collapses by $>$3 orders of magnitude (0.773 $\rightarrow$ 0.0002) within 12 steps, while KL to the training policy remains near zero throughout, indicating the model converges to a near-deterministic policy without drifting far from initialization. Group composition shifts from 81\% mixed groups at step 0 (productive gradient signal) to $>$96\% all-good groups by step 4 (saturated learning, diminishing returns).

\begin{table}[htbp]
    \centering
    \caption{Diagnostic telemetry from arithmetic GRPO training (Qwen3-8B, 100 steps, G=4). Entropy, KL, and group composition are logged per step; selected milestones shown.}
    \begin{tabular}{rcccccc}
        \toprule
        Step & Reward & Entropy & KL (v1) & Frac Mixed & Frac All Good & Frac All Bad \\
        \midrule
        0  & 0.676 & 0.773   & 5.7e-4  & 0.81 & 0.19 & 0.00 \\
        1  & 0.795 & 0.586   & 6.3e-4  & 0.62 & 0.37 & 0.01 \\
        2  & 0.933 & 0.228   & $-$1.4e-4 & 0.36 & 0.64 & 0.00 \\
        4  & 0.999 & 0.030   & 9.7e-5  & 0.04 & 0.96 & 0.00 \\
        7  & 0.990 & 0.004   & 3.0e-5  & 0.00 & 0.99 & 0.01 \\
        11 & 1.000 & 2.5e-4  & $-$6.6e-6 & 0.00 & 1.00 & 0.00 \\
        19 & 0.998 & 0.006   & $-$4.5e-5 & 0.01 & 0.99 & 0.00 \\
        \bottomrule
    \end{tabular}
    \label{tab:diagnostics}
\end{table}

\textbf{Key observations:}
\begin{itemize}
    \item \textbf{Entropy collapse:} 0.773 $\rightarrow$ 0.0002 (3,865$\times$ reduction) in $\sim$12 steps. Without explicit entropy regularization, the policy becomes near-deterministic, which limits further exploration.
    \item \textbf{KL remains near zero:} KL divergence to the initial training policy oscillates around $\pm$0.001 nats, indicating no KL anchor was needed for this short horizon / easy task. Longer or harder tasks may require explicit KL regularization.
    \item \textbf{Group saturation:} Once frac\_all\_good exceeds 0.95, within-group variance drops, advantages shrink, and effective learning rate diminishes (the ``zero-advantage'' regime). This provides a natural stopping criterion.
    \item \textbf{Missing for GSM8K/MATH:} Comparable entropy/KL/group-composition metrics were \emph{not logged} for the main GSM8K and MATH experiments (Tinker SDK did not expose these in the training loop). Adding this telemetry is a Phase 1 priority in our experimental roadmap (Section 7 of the main paper).
\end{itemize}

\subsection{MBPP Evaluation Failure}

The MBPP (Mostly Basic Python Problems) benchmark produced 0\% pass rate across all runs due to an output parsing bug in the evaluation script. The script expected a specific JSON format for extracted answers but the model output format differed, preventing correct answer matching.

This represents a methodological failure: an invalidated benchmark should not be reported. Only HumanEval results (which used a working evaluator) are reported in the main paper. Future work should verify evaluator correctness before running experiments.

% ===========================================================================
\section{Additional Experimental Details}
% ===========================================================================

\subsection{Reward Function Weighting Rationale}

The tool calling reward (0.3 + 0.4 + 0.3) was designed to prioritize:
\begin{itemize}
    \item \textbf{Correct tool name (+0.4):} Selecting the right tool is the core judgment decision
    \item \textbf{Valid JSON (+0.3):} Required for downstream parsing
    \item \textbf{Arguments present (+0.3):} Functional tool use requires populated arguments
\end{itemize}

These weights were not validated via ablation. Future work should test alternative weightings, particularly increasing the tool-name weight to further emphasize selection judgment over format compliance.

\subsection{Evaluation Protocol Limitations}

As noted in the main paper, our primary evaluation metrics are computed on \textbf{training prompts with new rollouts}, not on held-out test sets. Specifically:
\begin{itemize}
    \item Tool-use rewards: Fresh sampling from the same prompt distribution
    \item GSM8K accuracy: Random subset of 8 training examples per step (batch=2, group=4)
    \item HumanEval: Standard pass@1 on the 50-example benchmark (test set)
\end{itemize}

The GSM8K evaluation is the most compromised: evaluating on training data with small group sizes (G=4) means the ``test'' examples overlap with training distribution. True held-out evaluation on the GSM8K test set (1319 examples) was not performed.

\subsection{Multi-Turn Tool Calling Reward Breakdown}

The multi-turn reward function (5-component) enables finer-grained learning signals:

\begin{table}[htbp]
    \centering
    \caption{Multi-turn tool calling reward components.}
    \begin{tabular}{ll}
        \toprule
        Component & Weight \\
        \midrule
        First turn = tool call & +0.25 \\
        Final natural language answer & +0.30 \\
        Arguments populated & +0.15 \\
        Clean JSON output & +0.10 \\
        Repeated tool call (penalty) & -0.30 \\
        \bottomrule
    \end{tabular}
    \label{tab:multiturn_reward}
\end{table}

The -0.30 penalty for repeated tool calls was critical: it eliminated SFT's looping failure mode entirely (Section 4.1.2 of main paper). Without this penalty, models would repeatedly call the same tool in an infinite loop.

Because these scores come from custom scenario rollouts rather than a standard benchmark, we emphasize that they are internal protocol numbers. We did not measure inter-rater reliability, did not use multiple human judges, and did not adopt proxy-state final-answer evaluation. Future revisions should benchmark the same agents under ToolRM / FC-RewardBench-style outcome evaluation or other standardized multi-turn frameworks.

% ===========================================================================
\section{Qualitative Examples}
% ===========================================================================

\subsection{Successful Tool Call (Post-GRPO)}
\begin{verbatim}
<|im_start|>user
What is the weather in Bangalore?<|im_end|>
<|im_start|>assistant
{
  "tool_calls": [
    {
      "name": "get_weather",
      "arguments": {
        "location": "Bangalore",
        "unit": "celsius"
      }
    }
  ]
}<|im_end|>
\end{verbatim}

\subsection{Failed Tool Call (Pre-GRPO, Plain Text)}
\begin{verbatim}
<|im_start|>user
What is the weather in Bangalore?<|im_end|>
<|im_start|>assistant
The weather in Bangalore is sunny and warm today.
The temperature is around 28 degrees Celsius.
It should be a nice day to go outside!<|im_end|>
\end{verbatim}

The pre-GRPO model generates natural language responses instead of structured tool calls. After GRPO training, the model reliably produces valid JSON with correct tool names and populated arguments.

\subsection{Two-Phase Learning Trajectory (Representative)}

\begin{verbatim}
Step 1:  Reward=0.00, Loss=0.85, Output="42" (format: plain text)
Step 5:  Reward=0.05, Loss=0.80, Output="The answer is 42." (format: text with number)
Step 10: Reward=0.10, Loss=0.72, Output="\boxed{42}" (format: boxed answer)
Step 15: Reward=0.14, Loss=0.65, Output="\boxed{42}" (stable format)
Step 20: Reward=0.40, Loss=0.55, Output="\boxed{42}" (reasoning begins)
Step 25: Reward=0.58, Loss=0.40, Output="\boxed{42}" (reasoning improves)
Step 30: Reward=0.55, Loss=0.38, Output="\boxed{42}" (plateau)
\end{verbatim}

This trajectory illustrates the two-phase pattern: format compliance (steps 1-20) precedes reasoning improvement (steps 20-30).

% ===========================================================================
\section{Platform and Compute Costs}
% ===========================================================================

\begin{table}[htbp]
    \centering
    \caption{Compute platform usage and estimated costs.}
    \begin{tabular}{llcl}
        \toprule
        Platform & GPU & Experiments & Est. Cost \\
        \midrule
        Tinker SDK v0.16.1 & Cloud A100 & 17 runs & \$30--40 \\
        Google Colab Pro & T4 16GB & 5 team members & \$10/person \\
        HuggingFace Hub & N/A & Model hosting & Free \\
        \bottomrule
    \end{tabular}
    \label{tab:compute}
\end{table}

Total estimated spend: $\sim$\$50/person, well within typical capstone project budgets.

\begin{table}[htbp]
    \centering
    \caption{Estimated per-run compute budgets for representative experiments. Token estimates assume batch\_size=2, group\_size as configured, max\_generation=512 tokens. Wall-clock times are from logged telemetry where available; others are estimated from Tinker billing.}
    \begin{tabular}{llcccc}
        \toprule
        Domain & Config & Steps & Group & Est. Tokens & Est. Time \\
        \midrule
        Arithmetic & Qwen3-8B, r=32 & 100 & 4 & 410K & $\sim$17 min \\
        GSM8K & Qwen3-8B, r=32 & 50 & 8 & 410K & $\sim$25 min \\
        GSM8K & Qwen3-8B, r=32 & 100 & 8 & 820K & $\sim$50 min \\
        GSM8K & Qwen3.5-4B, r=32 & 50 & 8 & 410K & $\sim$20 min \\
        Tool (synthetic) & Qwen3-8B, r=32 & 30 & 8 & 245K & $\sim$15 min \\
        Tool (xlam-60k) & Qwen3-8B, r=32 & 100 & 8 & 820K & $\sim$50 min \\
        \bottomrule
    \end{tabular}
    \label{tab:per_run_budget}
\end{table}

\textbf{Limitations of budget reporting:} We do not report exact tokens processed per run (the Tinker SDK does not expose this metric). Token estimates above are upper bounds assuming maximum generation length. Wall-clock times exclude checkpoint saving overhead. GPU-hours are not tracked per-run; the aggregate across all 17 Tinker runs is estimated at 8--12 A100 GPU-hours.

We report GPU platform and wall-clock budget, but not a full token-processed accounting per run. Likewise, our data-split story is uneven across domains: HumanEval uses a held-out benchmark subset, while tool calling and GSM8K largely report training-distribution rollouts. This split mismatch is a major threat to cross-domain comparability.

% ===========================================================================
\section{Code and Reproducibility}
% ===========================================================================

All training scripts are available in the repository:
\begin{itemize}
    \item \texttt{grpo\_gsm8k\_base.py} -- Parameterized GSM8K training
    \item \texttt{grpo\_exp\_a\_baseline.py}, \texttt{exp\_b\_high\_lr.py}, \texttt{exp\_c\_low\_temp.py}, \texttt{exp\_d\_xlam.py} -- Tool-use experiments
    \item \texttt{grpo\_100\_xlam.py}, \texttt{grpo\_100\_synthetic.py}, \texttt{grpo\_100\_math.py} -- Extended runs
\end{itemize}

At present, we can release training scripts, paper sources, and the hardened held-out evaluation script, but the exact prompt packs / scenario definitions used for every internal tool-calling score are not yet packaged as a replication bundle. Releasing prompts, scoring rubrics, and scenario manifests is an explicit next step for reproducibility.

For stability diagnostics, we did not log KL-to-SFT trajectories or policy entropy, and our MoE analysis does not yet include routing entropy or expert-load imbalance measurements. Those omissions limit causal interpretation of both drift and volatility claims.

Representative training logs were collected during experimentation, but the literal local \texttt{/tmp} paths used during development are not a portable reproduction interface and should not be treated as a release artifact.

Checkpoint access is partial rather than universal at present:
\begin{itemize}
    \item Some trained artifacts were published to project-managed model hosting during development.
    \item Some tool-use and math checkpoints were written to Tinker under run-specific paths such as \texttt{tinker://\textit{run\_id}/sampler\_weights/final}, but present-day recovery depends on retention and account access.
\end{itemize}
A clean anonymous release should replace personal account identifiers with project-level artifact locations or explicit recovery instructions.

\end{document}
