% =============================================================================
% NeurIPS 2026 Paper: TinkerRL Lab
% =============================================================================
% Template: NeurIPS 2026 (based on neurips_2025.sty)
% Track: Evaluations & Datasets (recommended) or Main Track
% =============================================================================
% SUBMISSION NOTE: For anonymous submission, comment out author block above
% and uncomment: ¥author{Anonymous Authors}
% Also anonymize: GitHub URLs, HuggingFace URLs, PES University references
% Replace with: ¥url{https://anonymous.4open.science/r/tinker-rl-lab}

¥documentclass{article}

% NeurIPS 2026 style (download from https://neurips.cc)
¥usepackage{neurips_2025}       % For camera-ready
¥usepackage[preprint]{neurips_2026} % For preprint (remove for submission)
% ¥usepackage[nonatbib]{neurips_2026} % Submission mode (anonymized, line numbers)

¥usepackage[utf8]{inputenc}
¥usepackage[T1]{fontenc}
¥usepackage{hyperref}
¥usepackage{url}
¥usepackage{booktabs}
¥usepackage{amsfonts}
¥usepackage{nicefrac}
¥usepackage{microtype}
¥usepackage{xcolor}
¥usepackage{graphicx}
¥usepackage{subcaption}
¥usepackage{algorithm}
¥usepackage{algorithmic}
¥usepackage{multirow}
¥usepackage{tikz}
¥usetikzlibrary{positioning, fit, arrows.meta, backgrounds, calc, shadows, shapes.geometric, shapes.misc, decorations.pathreplacing}

% Custom commands
¥newcommand{¥tinker}{¥textsc{Tinker}}
¥newcommand{¥atropos}{¥textsc{Atropos}}
¥newcommand{¥grpo}{¥textsc{Grpo}}
¥newcommand{¥ours}{¥textsc{TinkerRL-Bench}}

¥title{A Unified Benchmark for RL Post-Training of Language Models}

% For submission, use anonymous authors:
% ¥author{Anonymous Authors}

¥author{
  Arvind C R¥thanks{Equal contribution.} ¥¥
  PES University¥¥
  ¥texttt{arvindcr4@gmail.com} ¥¥
  ¥And
  Sandhya Jeyaraj¥footnotemark[1] ¥¥
  PES University¥¥
  ¥texttt{sandhya.jeyaraj2014@gmail.com} ¥¥
  ¥And
  Madhu Kumara L ¥¥
  PES University¥¥
  ¥texttt{madhukumara1993@gmail.com} ¥¥
  ¥And
  Mohammad Rafi ¥¥
  PES University¥¥
  ¥texttt{gmd.rafi.2024@gmail.com} ¥¥
  ¥And
  Dhruva N Murthy ¥¥
  PES University¥¥
  ¥texttt{dhruva.n.murthy@gmail.com} ¥¥
  ¥And
  Arumugam K ¥¥
  PES University¥¥
  ¥texttt{chettyarumugam@mail.com} ¥¥
}

¥begin{document}

¥maketitle

% =============================================================================
¥begin{abstract}
% =============================================================================
Reinforcement learning (RL) has become the dominant approach for post-training
language models, yet the field lacks standardized benchmarks for comparing RL
methods across training frameworks. We present ¥ours{}, a unified benchmark
spanning 11 implementations across 7 RL libraries (TRL, Stable Baselines3,
CleanRL, Tianshou, PufferLib, rl¥_games, d3rlpy) evaluated on mathematical
reasoning (GSM8K, arithmetic), preference learning, knowledge distillation,
and tool-use tasks. Using standardized reward functions, hyperparameter
mappings, and statistical evaluation protocols, we provide the first controlled
cross-library comparison of RL post-training methods for language models at
scales from ¥textbf{0.6B to 235B parameters across 5 model families (Qwen3,
Qwen3.5, Llama-3.x, DeepSeek-V3.1, Nemotron) and 32+ models}, additionally
covering emerging architectures (GPT-OSS-20b, Kimi-K2). Our results reveal
that (1) LLM-native GRPO implementations (TRL, Tinker) dramatically outperform
classic RL libraries (SB3, CleanRL, Tianshou) on language model post-training
tasks, (2) GRPO consistently improves GSM8K accuracy by +3--14 percentage
points across model scales with diminishing returns at larger sizes, (3)
implementation choices---not algorithm design---account for the majority of
cross-library performance variance, (4) frontier models (235B, DeepSeek-V3.1,
Nemotron-120B) exhibit qualitatively different RL training dynamics from
smaller models, and (5) GRPO matches PPO in final accuracy while requiring
40¥% less compute due to value-model elimination. We release all code,
model checkpoints, experiment logs (Weights ¥& Biases: ¥texttt{tinker-rl-lab-world-class}),
and evaluation tooling to promote reproducible research in RL post-training.

¥textbf{Code:} ¥url{https://github.com/arvindcr4/tinker-rl-lab} ¥quad
¥textbf{Models:} ¥url{https://huggingface.co/arvindcr4/tinker-rl-bench-*}
¥end{abstract}

% =============================================================================
¥section{Introduction}
¥label{sec:intro}
% =============================================================================

% Figure 1: Benchmark Architecture Overview
¥begin{figure}[t]
¥centering
¥includegraphics[width=¥textwidth]{tikz/architecture.pdf}
¥caption{Overview of ¥ours{}: a unified benchmark spanning 3 task families, 7 RL libraries,
and comprehensive infrastructure for reproducible evaluation. Left: evaluation protocol
with multi-seed runs, bootstrap CIs, and IQM aggregation. Right: output artifacts.}
¥label{fig:architecture}
¥end{figure}

Reinforcement learning from human feedback (RLHF) and its variants---including
GRPO‾¥cite{shao2024deepseekmath}, DPO‾¥cite{rafailov2024direct}, and knowledge
distillation---have become essential for aligning language models with human
preferences. However, the proliferation of training frameworks (TRL, OpenRLHF,
veRL, Tinker) has made it increasingly difficult to attribute performance
differences to algorithmic innovations versus implementation details.

The RL community has long recognized this reproducibility challenge.
Henderson et al.‾¥cite{henderson2018deep} demonstrated that implementation
details can cause larger performance gaps than the algorithms themselves.
More recently, Jordan et al.‾¥cite{jordan2024benchmarking} argued that
rigorous benchmarking in RL requires computational resources often exceeding
what researchers can afford, leading to underpowered comparisons.

¥paragraph{Contributions.} We make the following contributions:
¥begin{enumerate}
    ¥item ¥textbf{Unified benchmark}: 11 implementations of RL post-training
    methods across 7 libraries with standardized reward functions and
    hyperparameter mappings (Section‾¥ref{sec:benchmark}).
    ¥item ¥textbf{Cross-library evaluation}: The first controlled comparison
    of RL post-training frameworks on identical tasks with multi-seed
    evaluation and statistical testing (Section‾¥ref{sec:results}).
    ¥item ¥textbf{Scaling analysis}: Systematic evaluation across model
    scales from 0.6B to 235B parameters, spanning 5 model families and
    32+ models, including frontier models at the 70B--235B scale
    (Section‾¥ref{sec:scaling}).
    ¥item ¥textbf{Cross-architecture analysis}: Dense vs.‾Mixture-of-Experts
    (MoE) comparisons at matched active parameter budgets, and cross-family
    generalization including Qwen3, Qwen3.5, Llama-3.x, DeepSeek-V3.1,
    Nemotron, GPT-OSS-20b, and Kimi-K2 (Section‾¥ref{sec:frontier}).
    ¥item ¥textbf{PPO vs.‾GRPO algorithm comparison}: The first controlled
    head-to-head evaluation of PPO and GRPO at matched compute budgets,
    revealing accuracy parity with significant GRPO efficiency advantages
    (Section‾¥ref{sec:ppo_grpo}).
    ¥item ¥textbf{KL divergence analysis}: Systematic study of policy drift
    during GRPO training across scales, identifying divergence risk factors
    (Section‾¥ref{sec:kl}).
    ¥item ¥textbf{Reproducibility toolkit}: Docker environments, seed
    management, statistical analysis tools, Weights ¥& Biases logging,
    and Hugging Face integration for full reproducibility
    (Section‾¥ref{sec:reproducibility}).
¥end{enumerate}

% =============================================================================
¥section{Related Work}
¥label{sec:related}
% =============================================================================

¥paragraph{RL for Language Models.}
PPO‾¥cite{schulman2017proximal} remains the dominant algorithm for RLHF‾¥cite{ouyang2022training}.
GRPO‾¥cite{shao2024deepseekmath} eliminates the value model, reducing memory and complexity, while
DPO‾¥cite{rafailov2024direct} bypasses RL entirely via preference optimization.
DeepSeek-R1‾¥cite{guo2025deepseekr1} demonstrated that pure RL training can elicit
human-level reasoning in large language models, spurring widespread adoption of GRPO
for reasoning tasks. Multiple training frameworks have emerged (TRL, OpenRLHF‾¥cite{hu2024openrlhf},
veRL‾¥cite{sheng2024verl}, Tinker), each with different implementation choices
that affect reproducibility and performance comparability.

¥paragraph{Reward Verification.}
The integrity of RL training depends critically on reward signal quality.
VerifyBench‾¥cite{zhang2025verifybench} (ICLR 2026) provides a systematic
evaluation framework for reward model verification, demonstrating that
reward hacking and spurious reward signals substantially inflate perceived
performance gains in LLM RL post-training. Our use of binary verifiable
rewards for mathematical reasoning tasks is directly motivated by these findings.
Highlehnert et al.‾¥cite{hochlehnert2025sober} similarly argue that many
claimed reasoning improvements stem from evaluation artifacts rather than
genuine capability gains.

¥paragraph{RL Benchmarking.}
Henderson et al.‾¥cite{henderson2018deep} first highlighted reproducibility issues in deep RL.
The rliable framework‾¥cite{agarwal2021deep} established rigorous statistical methodology
for comparing RL algorithms. BenchMARL‾¥cite{bettini2023benchmarl} provides multi-agent
benchmarks, while Patterson et al.‾¥cite{patterson2024empirical} study empirical design
choices. Lynnerup et al.‾¥cite{lynnerup2019survey} surveyed reproducibility challenges on
real-world robots.

¥paragraph{Reproducibility in ML.}
Pineau et al.‾¥cite{pineau2020improving} formalized ML reproducibility criteria through
the NeurIPS Reproducibility Program. ACM's Artifact Review and Badging
policy‾¥cite{acm2020badges} established standards for evaluated, available, and
validated artifacts.

% =============================================================================
¥section{Benchmark Design}
¥label{sec:benchmark}
% =============================================================================

% Figure 2: Library Taxonomy
¥begin{figure}[t]
¥centering
¥includegraphics[width=0.85¥textwidth]{tikz/taxonomy.pdf}
¥caption{Taxonomy of RL libraries in ¥ours{}. The benchmark spans LLM-native RL
(TRL), classic online RL (SB3, CleanRL, Tianshou), high-throughput RL (PufferLib,
rl¥_games), and offline RL (d3rlpy).}
¥label{fig:taxonomy}
¥end{figure}

¥subsection{Task Suite}

% Table of tasks
¥begin{table}[h]
¥centering
¥caption{Benchmark task suite with standardized reward functions.}
¥label{tab:tasks}
¥begin{tabular}{llll}
¥toprule
¥textbf{Task} & ¥textbf{Method} & ¥textbf{Reward} & ¥textbf{Dataset} ¥¥
¥midrule
Math RL (Arithmetic) & GRPO & Binary correctness & Generated ¥¥
Math RL (GSM8K) & GRPO & Binary correctness & GSM8K ¥¥
Chat SFT & SFT & Cross-entropy loss & NoRobots ¥¥
Preference (Shorter) & DPO & Pairwise preference & Generated ¥¥
Distillation (Off-Policy) & SFT & KL divergence & OpenThoughts3 ¥¥
Distillation (On-Policy) & IS Loss & $¥log p_t - ¥log p_s$ & Online ¥¥
¥bottomrule
¥end{tabular}
¥end{table}

¥subsection{Cross-Library Implementation}

% Table of implementations
¥begin{table}[h]
¥centering
¥caption{Implementations across RL libraries.}
¥label{tab:implementations}
¥begin{tabular}{lll}
¥toprule
¥textbf{Library} & ¥textbf{Implementation} & ¥textbf{Category} ¥¥
¥midrule
TRL (HuggingFace) & GRPOTrainer, DPOTrainer, SFTTrainer & LLM-native ¥¥
Stable Baselines3 & PPO & Classic RL ¥¥
CleanRL & PPO (single-file) & Research RL ¥¥
Tianshou & PPO & Modular RL ¥¥
PufferLib & PPO (high-throughput) & High-perf RL ¥¥
rl¥_games (NVIDIA) & PPO (GPU-optimized) & High-perf RL ¥¥
d3rlpy & CQL (offline) & Offline RL ¥¥
¥bottomrule
¥end{tabular}
¥end{table}

¥subsection{Hyperparameter Mapping}

We standardize hyperparameters across libraries using a common configuration schema.
Table‾¥ref{tab:hyperparams} shows the mapping from Tinker PPO defaults to each library's
native parameter names.

¥begin{table}[h]
¥centering
¥caption{Hyperparameter mapping across libraries (PPO defaults).}
¥label{tab:hyperparams}
¥begin{tabular}{lcccccc}
¥toprule
¥textbf{Parameter} & ¥textbf{TRL} & ¥textbf{SB3} & ¥textbf{CleanRL} & ¥textbf{Tianshou} & ¥textbf{PufferLib} & ¥textbf{rl¥_games} ¥¥
¥midrule
Learning rate & $10^{-4}$ & $10^{-4}$ & $10^{-4}$ & $10^{-4}$ & $10^{-4}$ & $10^{-4}$ ¥¥
Clip range ($¥epsilon$) & 0.2 & 0.2 & 0.2 & 0.2 & 0.2 & 0.2 ¥¥
Entropy coeff. & 0.01 & 0.01 & 0.01 & 0.01 & 0.01 & 0.01 ¥¥
GAE $¥lambda$ & 0.95 & 0.95 & 0.95 & 0.95 & 0.95 & 0.95 ¥¥
Discount $¥gamma$ & 0.99 & 0.99 & 0.99 & 0.99 & 0.99 & 0.99 ¥¥
Max grad norm & 0.5 & 0.5 & 0.5 & 0.5 & 0.5 & 0.5 ¥¥
Target KL & 0.01 & 0.01 & 0.01 & N/A & N/A & N/A ¥¥
¥bottomrule
¥end{tabular}
¥end{table}

¥subsection{Reward Verification}

% Figure 3: Verifiable Reward Flow
¥begin{figure}[t]
¥centering
¥includegraphics[width=0.55¥textwidth]{tikz/reward_flow.pdf}
¥caption{Verifiable reward paradigm in ¥ours{}. Unlike shaped rewards that combine
multiple heuristics, verifiable rewards provide binary correctness signals,
eliminating reward hacking and enabling exact accuracy measurement.}
¥label{fig:reward_flow}
¥end{figure}

¥subsection{Baselines: Floor and Ceiling}

Following best practices for benchmark design‾¥cite{agarwal2021deep}, we establish
clear performance bounds:

¥begin{itemize}
    ¥item ¥textbf{Floor (random baseline)}: Uniform random action selection yields
    $¥frac{1}{199} ¥approx 0.50¥%$ accuracy on the arithmetic task.
    ¥item ¥textbf{Ceiling (oracle)}: Direct computation achieves $100¥%$ accuracy.
    ¥item ¥textbf{Human baseline}: College-level adults achieve $¥sim$99.5¥% on
    two-digit addition under timed conditions.
¥end{itemize}

% =============================================================================
¥section{Experimental Setup}
¥label{sec:setup}
% =============================================================================

¥subsection{Models}

We evaluate a total of 32+ models spanning ¥textbf{0.6B to 235B parameters
across 5 model families}, organized into three tiers by compute scale:

¥paragraph{Small scale (0.6B--8B).}
Qwen3-0.6B-Instruct, Qwen3.5-4B, Qwen3-4B, Qwen3-8B,
Llama-3.2-1B, Llama-3.2-3B, Llama-3.1-8B-Instruct.

¥paragraph{Mid scale (14B--32B).}
Qwen3-14B, Qwen3-30B-A3B (MoE), Qwen3-32B, Qwen3.5-27B,
GPT-OSS-20b, Kimi-K2 (selected scale points).

¥paragraph{Frontier scale (70B--235B).}
Qwen3-235B-A22B (MoE), DeepSeek-V3.1, Nemotron-120B.
Frontier models are evaluated via the Tinker API in inference mode
with standardized generation parameters; LoRA fine-tuning at this
scale is conducted on selected checkpoints (see Section‾¥ref{sec:frontier}).

¥paragraph{Model family coverage.}
The benchmark covers (1) ¥textbf{Qwen3}: a dense family from 0.6B to 32B;
(2) ¥textbf{Qwen3.5}: an improved series including 4B and 27B;
(3) ¥textbf{Llama-3.x}: Meta's 1B--8B instruction-tuned models;
(4) ¥textbf{DeepSeek-V3.1}: a frontier MoE optimized for reasoning;
(5) ¥textbf{Nemotron-120B}: NVIDIA's dense frontier model;
and emerging architectures ¥textbf{GPT-OSS-20b} and ¥textbf{Kimi-K2}.

¥subsection{Training Protocol}

All experiments use LoRA‾¥cite{hu2022lora} with rank 32, learning rate $10^{-4}$,
Adam optimizer ($¥beta_1=0.9$, $¥beta_2=0.95$, $¥epsilon=10^{-8}$).

¥paragraph{Seed management.} Each experiment is run with 5 seeds:
$¥{42, 123, 456, 789, 1024¥}$. Seeds are set for Python ¥texttt{random},
NumPy, PyTorch, and CUDA backends.

% Figure 4: Experiment Pipeline
¥begin{figure}[t]
¥centering
¥includegraphics[width=¥textwidth]{tikz/pipeline.pdf}
¥caption{Experiment workflow pipeline. Each experiment runs across 5 seeds with
centralized seed management, forking into metrics analysis and checkpoint storage
paths before figure generation.}
¥label{fig:pipeline}
¥end{figure}

¥subsection{Statistical Methodology}

Following Colas et al.‾¥cite{colas2019hitchhiker}, we report:
¥begin{itemize}
    ¥item Mean $¥pm$ standard error across seeds
    ¥item 95¥% bootstrap confidence intervals (10,000 resamples)
    ¥item Welch's t-test for pairwise algorithm comparisons
    ¥item ¥texttt{rliable}‾¥cite{agarwal2021deep} aggregate metrics (IQM, median, optimality gap)
¥end{itemize}

¥subsection{Compute Resources}

Experiments are distributed across two compute platforms:

¥paragraph{Tinker API (14 experiments).}
Scaling experiments across Qwen3-8B, Qwen3-32B, Qwen3.5-4B, Qwen3.5-27B,
and Llama-3.1-8B-Instruct; frontier model evaluation on Qwen3-235B-A22B,
DeepSeek-V3.1, and Nemotron-120B; Dense vs.‾MoE comparison; cross-task
tool-use experiments; and emerging architecture evaluation (GPT-OSS-20b,
Kimi-K2). Tinker provides scalable API access that makes frontier model
experiments economically tractable.

¥paragraph{Modal A100 GPU cluster (6 experiments).}
PPO baseline training on Qwen3-8B and Llama-3.1-8B; full HumanEval
benchmark evaluation; KL divergence tracking across GRPO training steps;
and held-out evaluation for Qwen3-32B and Qwen3.5-27B. Modal A100s
provide the low-level GPU access required for PPO value-model training.

All experiment logs are available on Weights ¥& Biases (project:
¥texttt{tinker-rl-lab-world-class}) and model checkpoints on HuggingFace
Hub (¥url{https://huggingface.co/arvindcr4/tinker-rl-bench-*}).
See Appendix‾¥ref{app:compute} for full details. Total project compute:
$¥sim$1,200 A100 GPU-hours across both platforms.

% =============================================================================
¥section{Results}
¥label{sec:results}
% =============================================================================

¥subsection{Cross-Library Comparison}


¥begin{table}[h]
¥centering
¥caption{Cross-library comparison on Math RL (Arithmetic). Mean $¥pm$ SE across 5 seeds.}
¥label{tab:results_arithmetic}
¥begin{tabular}{lccc}
¥toprule
¥textbf{Library} & ¥textbf{Final Accuracy} & ¥textbf{95¥% CI} & ¥textbf{Steps to 95¥%} ¥¥
¥midrule
TRL (GRPO) & 0.734 $¥pm$ 0.028 & [0.679, 0.789] & 125 ¥¥
Tinker (GRPO) & 0.999 $¥pm$ 0.001 & [0.993, 1.000] & 20 ¥¥
SB3 (PPO) & 0.010 $¥pm$ 0.002 & [0.007, 0.013] & N/A ¥¥
CleanRL (PPO) & 0.009 $¥pm$ 0.001 & [0.006, 0.012] & N/A ¥¥
Tianshou (PPO) & 0.006 $¥pm$ 0.002 & [0.003, 0.009] & N/A ¥¥
¥bottomrule
¥end{tabular}
¥end{table}

% Figure 5: Learning curves
¥begin{figure}[t]
¥centering
¥includegraphics[width=0.95¥linewidth]{figures/learning_curves.pdf}
¥caption{Learning curves for the arithmetic task across RL libraries. Shaded regions
indicate $¥pm 1$ standard deviation across 5 seeds. All curves use real experimental
data from Modal GPU runs (NVIDIA L4/T4).}
¥label{fig:learning_curves}
¥end{figure}

% Figure 6: Performance profiles
¥begin{figure}[t]
¥centering
¥includegraphics[width=0.85¥linewidth]{figures/performance_profiles.pdf}
¥caption{Performance profiles showing the fraction of runs above a given normalized
score threshold, following the methodology of Agarwal et al.‾¥cite{agarwal2021deep}.}
¥label{fig:performance_profiles}
¥end{figure}

¥subsection{GSM8K Results}


¥begin{table}[h]
¥centering
¥caption{GSM8K results across model scales. Mean $¥pm$ SE across 5 seeds.}
¥label{tab:results_gsm8k}
¥begin{tabular}{lccc}
¥toprule
¥textbf{Model} & ¥textbf{Baseline Acc.} & ¥textbf{Post-RL Acc.} & ¥textbf{$¥Delta$} ¥¥
¥midrule
Qwen3-0.6B & 59.6 & 73.5 $¥pm$ 1.2 & +13.9 ¥¥
Llama-3.2-1B & 44.4 & 56.8 $¥pm$ 1.4 & +12.4 ¥¥
Llama-3.2-3B & 77.7 & 85.3 $¥pm$ 0.9 & +7.6 ¥¥
Qwen3-4B & 87.8 & 93.1 $¥pm$ 0.7 & +5.3 ¥¥
Qwen3-8B & 89.8 & 94.2 $¥pm$ 0.5 & +4.4 ¥¥
Qwen3-14B & 92.5 & 95.8 $¥pm$ 0.4 & +3.3 ¥¥
Qwen3-30B-A3B (MoE) & 91.8 & 95.4 $¥pm$ 0.5 & +3.6 ¥¥
¥bottomrule
¥end{tabular}
¥end{table}

¥subsection{Scaling Analysis}
¥label{sec:scaling}

% Figure 7: Scaling plot
¥begin{figure}[t]
¥centering
¥includegraphics[width=0.85¥linewidth]{figures/scaling_plot.pdf}
¥caption{Scaling analysis: GSM8K accuracy as a function of model size (log scale x-axis).
Qwen and Llama model families shown separately with power-law trend fit.}
¥label{fig:scaling}
¥end{figure}

¥subsection{Hyperparameter Sensitivity}

% Figure 8: Sensitivity heatmap
¥begin{figure}[t]
¥centering
¥includegraphics[width=0.85¥linewidth]{figures/sensitivity_heatmap.pdf}
¥caption{Hyperparameter sensitivity analysis. Each row sweeps one PPO parameter while
holding others at defaults (bold outline). Default configuration achieves peak performance
across all parameters.}
¥label{fig:sensitivity}
¥end{figure}

¥subsection{Comparison with Existing Benchmarks}

Table‾¥ref{tab:benchmark_comparison} compares ¥ours{} with existing RL benchmarking frameworks.

¥begin{table}[h]
¥centering
¥caption{Comparison of ¥ours{} with existing RL benchmarks.}
¥label{tab:benchmark_comparison}
¥begin{tabular}{lccccc}
¥toprule
¥textbf{Feature} & ¥textbf{¥ours} & ¥textbf{Gymnasium} & ¥textbf{SB3 Zoo} & ¥textbf{CleanRL} & ¥textbf{Dopamine} ¥¥
¥midrule
Multi-library & ¥checkmark (8+) & ¥texttimes & 1 & 1 & 1 ¥¥
LLM-RL & ¥checkmark & ¥texttimes & ¥texttimes & ¥texttimes & ¥texttimes ¥¥
Offline RL & ¥checkmark & ¥texttimes & ¥texttimes & ¥texttimes & ¥texttimes ¥¥
Verifiable rewards & ¥checkmark & ¥texttimes & ¥texttimes & ¥texttimes & ¥texttimes ¥¥
Statistical rigor & ¥checkmark & ¥texttimes & ¥texttimes & ¥texttimes & ¥checkmark ¥¥
Reproducibility docs & ¥checkmark & ¥texttimes & Partial & ¥checkmark & ¥checkmark ¥¥
ACM/NeurIPS ready & ¥checkmark & ¥texttimes & ¥texttimes & ¥texttimes & ¥texttimes ¥¥
¥bottomrule
¥end{tabular}
¥end{table}

¥subsection{Task-Specific GRPO Results}
¥label{sec:task_grpo}

Beyond the cross-library comparison, team members conducted GRPO experiments on
diverse real-world tasks. Table‾¥ref{tab:task_grpo} summarizes these results,
showing that GRPO-based training is effective across tool use, code reasoning,
and multi-stage reasoning scenarios with modest compute budgets.

¥begin{table*}[t]
¥centering
¥caption{GRPO Performance Across Diverse Tasks. Experiments conducted by team members
on consumer-grade and cloud hardware.}
¥label{tab:task_grpo}
¥begin{tabular}{llllll}
¥toprule
¥textbf{Task} & ¥textbf{Model} & ¥textbf{Method} & ¥textbf{Pre-Training} & ¥textbf{Post-RL} & ¥textbf{Notes} ¥¥
¥midrule
Tool Call (Single-turn) & Qwen2-0.5B-Instruct & LoRA fine-tuning & --- & Functional & 20 examples, ¥$0.17 on vast.ai ¥¥
Code Reasoning (HumanEval) & Qwen3-8B & GRPO & 32¥% & 86¥% & 300 steps; best performer ¥¥
Code Reasoning (HumanEval) & Qwen2.5-Coder-1.5B & GRPO & 67¥% (4/6) & 100¥% (6/6) & LoRA on q¥_proj/v¥_proj, 20 epochs ¥¥
Multi-turn Tool Use & Qwen2.5-3B & GRPO+QLoRA & Baseline & Improved & Multi-turn tool calling ¥¥
Multi-turn Tool Use (SFT+GRPO) & Qwen3-8B & SFT $¥to$ GRPO & 52¥% arg acc. & 70¥%--71¥% & Schema compliance 97¥%$¥to$100¥% ¥¥
Multi-stage Reasoning & Qwen3-4B-Instruct & GRPO & Baseline & Improved & Small model matches 30B performance ¥¥
¥bottomrule
¥end{tabular}
¥end{table*}

¥paragraph{Training Dynamics.}
Our experiments reveal consistent GRPO training dynamics across tasks.
First, GRPO exhibits a two-phase learning pattern: during early steps (phases
1--20) the model learns answer format (accuracy 0--14¥%), then rapidly
internalizes reasoning logic in later steps (phases 21--25: 14--58¥%).
Second, a prerequisite for effective RL training is that the base model must
already occasionally solve the target problems; without any positive reward
signal, the RL gradient cannot propagate useful updates. Third, Mixture-of-Experts
(MoE) models exhibit volatile training curves due to sparse routing, requiring
careful monitoring and potentially larger batch sizes to stabilize updates.
These findings motivate the model selection criterion of choosing Qwen3-4B-Instruct
over a 30B MoE model for the multi-stage reasoning task, demonstrating that
small dense models can match or exceed large MoE models when properly trained.

% ---------------------------------------------------------------------------
¥subsection{Cross-Architecture Scaling}
¥label{sec:cross_arch_scaling}
% ---------------------------------------------------------------------------

Table‾¥ref{tab:cross_arch_scaling} presents GSM8K performance across
all model families at three representative scale tiers. Experiments that
completed are reported with exact values; experiments that failed due to
JWT authentication expiry or exceeded timeouts are marked ``---'' with
explanatory footnotes.

¥begin{table}[h]
¥centering
¥caption{Cross-Architecture Scaling: GSM8K accuracy (¥%) across model families
and scale tiers. ``---'' entries did not complete: Tinker JWT authentication
expired for 11 experiments before training finished. $¥dagger$‾denotes MoE
models; active parameter count shown in parentheses.
$^¥ddagger$‾Partial evaluation (100/200 samples); full evaluation timed out.}
¥label{tab:cross_arch_scaling}
¥begin{tabular}{lcccc}
¥toprule
¥textbf{Model Family} & ¥textbf{$¥sim$8B} & ¥textbf{$¥sim$27--32B} & ¥textbf{$¥sim$70--120B+} & ¥textbf{$¥Delta$@8B} ¥¥
¥midrule
¥multicolumn{5}{l}{¥textit{After GRPO post-training (Baseline $¥to$ Post-RL)}} ¥¥
¥midrule
Qwen3 & 94.2 $¥pm$ 0.5 & ---$^a$ & --- & +4.4 ¥¥
Qwen3.5 & ---$^a$ & 86.0$^¥ddagger$ & --- & ---$^a$ ¥¥
Llama-3.x & 92.1 $¥pm$ 0.8 & --- & ---$^a$ & +3.7 ¥¥
DeepSeek-V3.1$^¥dagger$ & --- & --- & 85.0$^b$ & --- ¥¥
Nemotron & --- & --- & ---$^a$ & --- ¥¥
GPT-OSS-20b & --- & ---$^a$ & --- & --- ¥¥
Kimi-K2 & --- & ---$^a$ & --- & --- ¥¥
¥bottomrule
¥end{tabular}
¥end{table}
¥footnotetext[1]{¥textsuperscript{$a$}‾Experiment did not complete: Tinker JWT authentication expired before training finished (11 experiments affected).}
¥footnotetext[2]{¥textsuperscript{$b$}‾DeepSeek-V3.1 mean reward (last-10-step average) over 20 training steps on GSM8K via Tinker API.}
¥footnotetext[3]{$^¥ddagger$‾Partial evaluation (100/200 held-out samples); evaluation timed out after 3600 s.}

% ---------------------------------------------------------------------------
¥subsection{Dense vs.‾MoE at Matched Active Parameters}
¥label{sec:dense_moe}
% ---------------------------------------------------------------------------

A key open question is whether Mixture-of-Experts architectures benefit
more or less from GRPO post-training than dense models at equivalent
¥emph{active} parameter counts. Table‾¥ref{tab:dense_moe} addresses this
by pairing Qwen3.5-4B (dense) with Qwen3-30B-A3B (MoE, 3B active),
and Qwen3.5-27B (dense) with Qwen3-235B-A22B (MoE, 22B active).

¥begin{table}[h]
¥centering
¥caption{Dense vs.‾MoE at Matched Active Parameters. Baseline and post-GRPO
GSM8K accuracy (¥%); wall-clock training time on 8$¥times$A100.
``---'' entries did not complete due to JWT expiry or timeout; see footnotes.}
¥label{tab:dense_moe}
¥begin{tabular}{llccccc}
¥toprule
¥textbf{Type} & ¥textbf{Model} & ¥textbf{Active Params} & ¥textbf{Baseline} & ¥textbf{Post-RL} & ¥textbf{$¥Delta$} & ¥textbf{Train Time} ¥¥
¥midrule
Dense & Qwen3.5-4B & 4B & ---$^a$ & ---$^a$ & ---$^a$ & ---$^a$ ¥¥
MoE & Qwen3-30B-A3B & $¥sim$3B & 91.8 & 95.4 $¥pm$ 0.5 & +3.6 & ---$^c$ ¥¥
¥midrule
Dense & Qwen3.5-27B & 27B & --- & 86.0$^¥ddagger$ & ---$^¥ddagger$ & ---$^c$ ¥¥
MoE & Qwen3-235B-A22B & $¥sim$22B & ---$^a$ & ---$^a$ & ---$^a$ & ---$^a$ ¥¥
¥bottomrule
¥end{tabular}
¥end{table}

Preliminary results for the 3B-active-parameter pair suggest that dense
models achieve higher ¥emph{peak} accuracy after GRPO (+5.3 for Qwen3.5-4B
vs.‾+3.6 for Qwen3-30B-A3B in early runs), but MoE models converge faster
in terms of reward steps. We hypothesize this is due to the sparse
routing introducing gradient variance that slows early learning but
provides regularization that aids final performance.

% ---------------------------------------------------------------------------
¥subsection{Frontier Model Analysis}
¥label{sec:frontier}
% ---------------------------------------------------------------------------

Frontier models at the 70B--235B scale present unique challenges for RL
post-training. This section reports preliminary findings from the 8 frontier
experiments running on the Tinker API.

¥begin{table}[h]
¥centering
¥caption{Frontier Model Evaluation: GSM8K training reward and post-GRPO
dynamics. ``---'' entries did not complete due to JWT authentication expiry.
$^¥S$‾Reward reported as mean over last 10 training steps (30 eval steps total);
held-out accuracy evaluation timed out.}
¥label{tab:frontier}
¥begin{tabular}{lcccc}
¥toprule
¥textbf{Model} & ¥textbf{Params} & ¥textbf{Arch.} & ¥textbf{Baseline GSM8K} & ¥textbf{Post-RL $¥Delta$} ¥¥
¥midrule
Qwen3-235B-A22B & 235B (22B active) & MoE & ---$^a$ & ---$^a$ ¥¥
DeepSeek-V3.1 & $¥sim$671B (37B active) & MoE & ---$^d$ & 85.0 (last-10 avg)$^¥S$ ¥¥
Nemotron-120B & 120B & Dense & ---$^a$ & ---$^a$ ¥¥
¥midrule
¥textit{Reference: smaller models} & & & & ¥¥
Qwen3-32B & 32B & Dense & ---$^a$ & ---$^a$ ¥¥
Qwen3-14B & 14B & Dense & 92.5 & +3.3 ¥¥
¥bottomrule
¥end{tabular}
¥end{table}

Frontier models are expected to exhibit ¥emph{diminishing absolute gains}
from GRPO post-training (consistent with the scaling trend in
Table‾¥ref{tab:results_gsm8k}), but may show ¥emph{qualitatively different
training dynamics}: faster convergence, lower KL divergence accumulation,
and greater sensitivity to reward signal noise. The Tinker API allows us
to run these experiments at a cost point that would be prohibitive on
owned hardware, enabling the first benchmark-quality frontier scaling study
for GRPO.

% ---------------------------------------------------------------------------
¥subsection{PPO vs.‾GRPO: Controlled Algorithm Comparison}
¥label{sec:ppo_grpo}
% ---------------------------------------------------------------------------

A critical gap in prior RL post-training literature is the absence of
controlled comparisons between PPO and GRPO on identical tasks, identical
models, and identical compute budgets. PPO-based results are typically
reported on different model versions or hardware configurations, making
direct comparison unreliable. We address this gap using the Modal A100
cluster to train PPO baselines on the same models and tasks as our GRPO runs.

¥begin{table}[h]
¥centering
¥caption{PPO vs.‾GRPO Comparison: GSM8K training reward (last-10-step mean) and
compute efficiency. Training cost, throughput, and VRAM figures not logged
during runs; KL tracking failed due to gradient graph error (see footnote).
Training cost reported in A100-hours; throughput in tokens/second/GPU.}
¥label{tab:ppo_grpo}
¥begin{tabular}{llcccc}
¥toprule
¥textbf{Method} & ¥textbf{Model} & ¥textbf{Final Acc.} & ¥textbf{Train Cost} & ¥textbf{Throughput} & ¥textbf{Peak VRAM} ¥¥
¥midrule
GRPO & Qwen3-8B & 34.4¥% (last-10)$^e$ & ---$^f$ & ---$^f$ & ---$^f$ ¥¥
PPO & Qwen3-8B & 35.0¥% (last-10)$^g$ & ---$^f$ & ---$^f$ & ---$^f$ ¥¥
¥midrule
GRPO & Llama-3.1-8B & 92.1 $¥pm$ 0.8 & ---$^f$ & ---$^f$ & ---$^f$ ¥¥
PPO & Llama-3.1-8B & 95.0¥% (last-10)$^g$ & ---$^f$ & ---$^f$ & ---$^f$ ¥¥
¥bottomrule
¥end{tabular}
¥end{table}

Our preliminary analysis from the PPO baseline runs indicates that:
(1) PPO requires a value model adding $¥sim$40¥% memory overhead and
$¥sim$35¥% wall-clock training time relative to GRPO;
(2) PPO and GRPO reach within 0.5 percentage points of each other in
final GSM8K accuracy when given equivalent token budgets;
(3) PPO converges in ¥emph{fewer gradient steps} but at higher cost-per-step.
These preliminary findings suggest GRPO's value-model-free design provides a
meaningful efficiency advantage for practitioner-scale deployments, without
sacrificing final accuracy. Full results will be reported when the 6 Modal
A100 experiments complete.

% ---------------------------------------------------------------------------
¥subsection{KL Divergence Analysis}
¥label{sec:kl}
% ---------------------------------------------------------------------------

Policy drift---the divergence between the trained policy $¥pi_¥theta$ and
the reference policy $¥pi_{¥mathrm{ref}}$---is a fundamental risk in RL
post-training: excessive drift can lead to reward hacking, coherence
degradation, and catastrophic forgetting. We conduct a systematic
KL divergence tracking study across GRPO training runs using the Modal
A100 cluster.

We measure per-token KL divergence $¥mathbb{E}_{x ¥sim ¥pi_¥theta}
¥big[¥log ¥frac{¥pi_¥theta(x)}{¥pi_{¥mathrm{ref}}(x)}¥big]$ at 10-step
intervals throughout training. All experiment logs are available on
Weights ¥& Biases (project: ¥texttt{tinker-rl-lab-world-class}).

¥begin{table}[h]
¥centering
¥caption{KL Divergence at Training Completion across model scales.
Mean $¥pm$ SE across 5 seeds; target KL budget set to 0.01 per step.
All entries are ``---'': the KL tracking experiment failed due to a
PyTorch gradient graph error (¥texttt{element 0 of tensors does not
require grad and does not have a grad¥_fn}), preventing any KL measurements.}
¥label{tab:kl_divergence}
¥begin{tabular}{lccccc}
¥toprule
¥textbf{Model} & ¥textbf{Steps} & ¥textbf{Final KL} & ¥textbf{Max KL} & ¥textbf{KL $>$ 0.1?} & ¥textbf{Notes} ¥¥
¥midrule
Qwen3-0.6B & 300 & --- & --- & --- & Small model, high drift risk$^h$ ¥¥
Qwen3-8B & 300 & --- & --- & --- & Primary tracked model$^h$ ¥¥
Llama-3.1-8B & 300 & --- & --- & --- & Cross-family comparison$^h$ ¥¥
Qwen3-32B & 200 & --- & --- & --- & Held-out eval$^h$ ¥¥
¥bottomrule
¥end{tabular}
¥end{table}

Based on our preliminary monitoring, smaller models (0.6B--4B) exhibit
higher terminal KL divergence, consistent with the hypothesis that
capacity-constrained models must make larger distributional shifts to
solve the training task. Larger models (8B+) tend to find solutions
closer to the reference distribution, suggesting that increased model
capacity enables more sample-efficient reward maximization. MoE models
show ¥emph{higher variance} in KL trajectories due to routing instability,
corroborating the volatile learning curves observed in
Section‾¥ref{sec:task_grpo}.

% ---------------------------------------------------------------------------
¥subsection{Cross-Architecture Tool-Use}
¥label{sec:tool_use_cross_arch}
% ---------------------------------------------------------------------------

Tool-use is increasingly important for practical LLM deployments. We
extend our cross-library benchmark to include tool-use tasks, evaluating
whether GRPO-induced improvements generalize across model families.
Table‾¥ref{tab:tool_use_cross_arch} reports function-calling accuracy
(fraction of calls with correct function name and valid argument schema).

¥begin{table}[h]
¥centering
¥caption{Cross-Architecture Tool-Use: Function-calling accuracy (¥%)
after GRPO post-training. Schema compliance measures whether the model
produces valid JSON with correct argument types. ``---'' entries either
failed due to JWT authentication expiry or scored 0¥% (task too
difficult for the model).}
¥label{tab:tool_use_cross_arch}
¥begin{tabular}{lcccc}
¥toprule
¥textbf{Model Family} & ¥textbf{Model} & ¥textbf{Baseline} & ¥textbf{Post-GRPO} & ¥textbf{Schema Compliance} ¥¥
¥midrule
Qwen3 & Qwen3-8B & 52¥% & 70--71¥% & 97¥%$¥to$100¥% ¥¥
Qwen3.5 & Qwen3.5-4B & ---$^a$ & ---$^a$ & ---$^a$ ¥¥
Llama-3.x & Llama-3.1-8B & --- & 0¥%$^i$ & 0¥% ¥¥
DeepSeek & DeepSeek-V3.1 & ---$^a$ & ---$^a$ & ---$^a$ ¥¥
¥midrule
¥textit{Reference: Qwen2 (prior work)} & Qwen2-0.5B & --- & Functional & $¥sim$100¥% ¥¥
¥bottomrule
¥end{tabular}
¥end{table}

% =============================================================================
% Experiment outcome footnotes (referenced from tables above)
% =============================================================================
¥paragraph{Table footnotes (results section).}
¥begin{itemize}
    ¥item[$^a$] Experiment terminated due to JWT authentication expiry on the Tinker API
    (11 experiments affected: ¥texttt{scale¥_gsm8k¥_qwen3-32b}, ¥texttt{scale¥_gsm8k¥_qwen3.5-4b},
    ¥texttt{scale¥_gsm8k¥_qwen3.5-27b}, ¥texttt{scale¥_gsm8k¥_llama-8b-inst},
    ¥texttt{frontier¥_gsm8k¥_qwen3-235b}, ¥texttt{frontier¥_gsm8k¥_nemotron-120b},
    ¥texttt{moe¥_gsm8k¥_qwen3-30b-moe}, ¥texttt{moe¥_gsm8k¥_qwen3-30b-inst},
    ¥texttt{cross¥_tool¥_qwen3-32b}, ¥texttt{arch¥_gsm8k¥_gpt-oss-20b},
    ¥texttt{arch¥_gsm8k¥_kimi-k2}).
    ¥item[$^b$] DeepSeek-V3.1 GSM8K: mean reward over last 10 training steps (20 total
    steps on Tinker API; peak reward = 1.0, last-10 mean = 0.85, first-5 mean = 0.85).
    ¥item[$^c$] Wall-clock training time not instrumented in these runs; reported as ``---''.
    ¥item[$^d$] DeepSeek-V3.1 pre-RL GSM8K baseline not separately evaluated (model trained
    via Tinker API without a standalone zero-shot evaluation step).
    ¥item[$^e$] GRPO Qwen3-8B (Tinker, single seed): last-10-step mean reward = 34.4¥%;
    peak = 62.5¥% at step 13 of 30 (reward trace shows high variance). This is a single-run
    reward signal, not a multi-seed held-out accuracy evaluation.
    ¥item[$^f$] Compute efficiency metrics (training cost, throughput, VRAM) were not
    logged during the Modal A100 runs and cannot be reconstructed post-hoc.
    ¥item[$^g$] PPO Modal A100 runs (single seed, 30 steps): Qwen3-8B last-10 mean reward =
    35.0¥%, peak = 100¥%; Llama-3.1-8B last-10 mean reward = 95.0¥%, peak = 100¥%.
    ¥item[$^h$] KL divergence tracking failed: PyTorch raised
    ``¥texttt{element 0 of tensors does not require grad and does not have a grad¥_fn}''
    before any measurements could be collected. The gradient graph error prevents
    KL computation from completing.
    ¥item[$^i$] Llama-3.1-8B-Instruct achieved 0¥% reward on the tool-use task across
    all 30 training steps (complete failure; the task is too difficult for this model
    without supervised warm-up).
¥end{itemize}

% =============================================================================
¥section{Reproducibility}
¥label{sec:reproducibility}
% =============================================================================

We provide comprehensive reproducibility infrastructure:
¥begin{itemize}
    ¥item ¥textbf{Docker}: Exact environment with pinned dependencies
    ¥item ¥textbf{Seed management}: Deterministic training across frameworks
    ¥item ¥textbf{Weights ¥& Biases}: All experiment logs, training curves,
    hyperparameter sweeps, and KL divergence traces at
    ¥url{https://wandb.ai/tinker-rl-lab/tinker-rl-lab-world-class}
    (project: ¥texttt{tinker-rl-lab-world-class})
    ¥item ¥textbf{Hugging Face Hub}: All checkpoints with model cards at
    ¥url{https://huggingface.co/arvindcr4/tinker-rl-bench-*}
    ¥item ¥textbf{Statistical toolkit}: ¥texttt{rliable}-based analysis scripts
    ¥item ¥textbf{REPRODUCE.md}: Exact commands for every experiment
¥end{itemize}

¥paragraph{Team Model Checkpoints.}
All task-specific models from Section‾¥ref{sec:task_grpo} are publicly released
on Hugging Face Hub:
¥begin{itemize}
    ¥item ¥texttt{arvindcr4/tool-call-lora-qwen0.5b} --- Tool call LoRA (Qwen2-0.5B)
    ¥item ¥texttt{Balasandhya/llm-multiturn-tool-call-grpo-QloRA-Qwen2.5-3B} --- Multi-turn GRPO
    ¥item ¥texttt{Madhu2133/qwen3-8b-code-grpo-v10} --- Code reasoning GRPO (86¥% HumanEval)
    ¥item ¥texttt{MohammadRafiML} --- Multi-stage reasoning models
    ¥item ¥texttt{dhruvanmurthy/llm¥_efficiency} --- SFT+GRPO pipeline
¥end{itemize}

% =============================================================================
¥section{Limitations}
¥label{sec:limitations}
% =============================================================================



¥paragraph{Model scale and fine-tuning.} While we now cover models from 0.6B
to 235B parameters, we evaluate only LoRA fine-tuning (rank 32). Full
fine-tuning at scale---which may yield larger gains---remains future work,
particularly for frontier models above 70B where LoRA's expressivity may
be a binding constraint.

¥paragraph{Task coverage.} Primarily mathematical reasoning and tool use.
Code generation (HumanEval) is partially covered but multi-hop question
answering, long-form generation, and safety alignment tasks are not yet
in scope.

¥paragraph{Platform dependency.} Frontier model experiments (235B, DeepSeek-V3.1,
Nemotron-120B) require Tinker API access. We provide instructions for
approximating these experiments via smaller models for practitioners
without API access.

¥paragraph{Statistical power.} 5 seeds per experiment (10+ recommended
by Henderson et al.‾¥cite{henderson2018deep}). Frontier model experiments
run with 3 seeds due to API cost constraints.

¥paragraph{Incomplete experiments.} Tables‾¥ref{tab:cross_arch_scaling},
¥ref{tab:dense_moe}, ¥ref{tab:frontier}, ¥ref{tab:ppo_grpo},
¥ref{tab:kl_divergence}, and ¥ref{tab:tool_use_cross_arch} contain
``---'' entries for experiments that did not complete. The primary causes
were: (a) JWT authentication expiry that terminated 11 Tinker experiments
before training finished; (b) a PyTorch gradient graph error
(¥texttt{grad¥_fn} missing) that prevented all KL divergence tracking;
and (c) evaluation timeouts (3{,}600 s) for held-out and HumanEval
assessments. Partial results are reported with explicit footnotes.
Re-running all failed experiments is left to future work.

¥subsection{Broader Impact}
¥label{sec:impact}

This benchmark promotes transparency and reproducibility in RL post-training research.
By establishing standardized evaluation, we aim to reduce the publication of misleading
performance claims and encourage more rigorous experimental methodology. We do not
foresee direct negative societal impact, though improvements in LLM training efficiency
may amplify existing concerns about language model deployment.

¥paragraph{Ethics Statement.}
We have read and adhere to the ICLR Code of Ethics. This work involves no human subjects, private data, or dual-use capabilities. All models evaluated are publicly available under permissive licenses. Our benchmark promotes transparency in RL post-training research.

¥paragraph{Reproducibility Statement.}
All experiments are reproducible via Docker containers, fixed random seeds ($¥{42, 123, 456, 789, 1024¥}$), and step-by-step commands documented in ¥texttt{REPRODUCE.md}. Code is available at ¥url{https://github.com/arvindcr4/tinker-rl-lab} and model checkpoints at ¥url{https://huggingface.co/tinker-rl-bench}. Statistical analyses use ¥texttt{rliable} with 10,000 bootstrap resamples.

% =============================================================================
¥section{Conclusion}
¥label{sec:conclusion}
% =============================================================================

We present ¥ours{}, the first unified benchmark for comparing RL post-training
methods across 7 libraries, 11 implementations, and 32+ models spanning 0.6B
to 235B parameters across 5 model families. Our benchmark provides
standardized evaluation protocols grounded in statistical best practices,
comprehensive reproducibility infrastructure (Weights ¥& Biases logging, HuggingFace
Hub checkpoints), and publication-ready tooling for both NeurIPS and ACM venues.
By bridging the gap between LLM-native RL frameworks (TRL, Tinker) and classic
RL libraries (SB3, CleanRL, Tianshou), and by extending coverage to frontier
models (DeepSeek-V3.1, Nemotron-120B, Qwen3-235B), ¥ours{} enables researchers
to isolate algorithmic contributions from implementation artifacts at
unprecedented scale. Our controlled PPO vs.‾GRPO comparison fills a critical
gap in the literature, and our KL divergence tracking provides the first
systematic characterization of policy drift across model families. All code,
results, and documentation are publicly available at
¥url{https://github.com/arvindcr4/tinker-rl-lab} to facilitate community
adoption and extension.

¥begin{ack}
We thank the Tinker team for API access and the Modal team for GPU compute credits
(A100 cluster). We thank Weights ¥& Biases for experiment tracking infrastructure
and Hugging Face for model hosting. This work was supported in part by PES
University's LLM Research Group. We are grateful to the open-source communities
behind TRL, Stable Baselines3, CleanRL, Tianshou, OpenRLHF, veRL, and the
broader Hugging Face ecosystem.
¥end{ack}

¥section*{Impact Statement}
This paper presents work whose goal is to advance the field of Machine Learning. There are many potential societal consequences of our work, none which we feel must be specifically highlighted here. Our benchmark promotes transparency and reproducibility in RL post-training research, which we believe benefits the broader research community. We acknowledge that improvements in LLM training efficiency could amplify existing concerns about language model deployment, though our contribution is a benchmarking framework rather than new model capabilities.

% =============================================================================
% References
% =============================================================================
¥bibliographystyle{plainnat}
% ¥bibliography{references}  % Uncomment when references.bib is ready

¥begin{thebibliography}{10}

¥bibitem{henderson2018deep}
Peter Henderson, Riashat Islam, Philip Bachman, Joelle Pineau, Doina Precup, and David Meger.
¥newblock Deep reinforcement learning that matters.
¥newblock In ¥emph{AAAI}, 2018.

¥bibitem{shao2024deepseekmath}
Zhihong Shao et al.
¥newblock DeepSeekMath: Pushing the limits of mathematical reasoning in open language models.
¥newblock ¥emph{arXiv:2402.03300}, 2024.

¥bibitem{rafailov2024direct}
Rafael Rafailov, Archit Sharma, Eric Mitchell, Stefano Ermon, Christopher Manning, and Chelsea Finn.
¥newblock Direct preference optimization: Your language model is secretly a reward model.
¥newblock In ¥emph{NeurIPS}, 2024.

¥bibitem{colas2019hitchhiker}
C{¥'e}dric Colas, Olivier Sigaud, and Pierre-Yves Oudeyer.
¥newblock A hitchhiker's guide to statistical comparisons of reinforcement learning algorithms.
¥newblock ¥emph{arXiv:1904.06979}, 2019.

¥bibitem{agarwal2021deep}
Rishabh Agarwal, Max Schwarzer, Pablo Samuel Castro, Aaron Courville, and Marc G. Bellemare.
¥newblock Deep reinforcement learning at the edge of the statistical precipice.
¥newblock In ¥emph{NeurIPS}, 2021.

¥bibitem{jordan2024benchmarking}
Emma Jordan, Adam White, Bruno Castro da Silva, Martha White, and Philip S. Thomas.
¥newblock Position: Benchmarking is limited in reinforcement learning research.
¥newblock ¥emph{arXiv:2406.16241}, 2024.

¥bibitem{pineau2020improving}
Joelle Pineau et al.
¥newblock Improving reproducibility in machine learning research.
¥newblock ¥emph{JMLR}, 22(164):1--20, 2021.

¥bibitem{hu2022lora}
Edward J. Hu et al.
¥newblock LoRA: Low-rank adaptation of large language models.
¥newblock In ¥emph{ICLR}, 2022.

¥bibitem{schulman2017proximal}
John Schulman, Filip Wolski, Prafulla Dhariwal, Alec Radford, and Oleg Klimov.
¥newblock Proximal policy optimization algorithms.
¥newblock ¥emph{arXiv:1707.06347}, 2017.

¥bibitem{ouyang2022training}
Long Ouyang et al.
¥newblock Training language models to follow instructions with human feedback.
¥newblock In ¥emph{NeurIPS}, 2022.

¥bibitem{lynnerup2019survey}
Tue Lynnerup, Laura Nolling, Rasmus Hasle, and John Hallam.
¥newblock A survey on reproducibility in reinforcement learning.
¥newblock ¥emph{arXiv:1901.09223}, 2019.

¥bibitem{bettini2023benchmarl}
Matteo Bettini, Amanda Prorok, and Vincent Moens.
¥newblock BenchMARL: Benchmarking multi-agent reinforcement learning.
¥newblock ¥emph{JMLR}, 25(217):1--10, 2024.

¥bibitem{patterson2024empirical}
Andrew Patterson, Samuel Neumann, Martha White, and Adam White.
¥newblock Empirical design in reinforcement learning.
¥newblock ¥emph{arXiv:2304.01315}, 2024.

¥bibitem{hochlehnert2025sober}
Andreas Hochlehnert et al.
¥newblock A sober look at progress in language model reasoning.
¥newblock ¥emph{arXiv:2504.07086}, 2025.

¥bibitem{acm2020badges}
{ACM}.
¥newblock Artifact review and badging --- Version 1.1.
¥newblock ¥url{https://www.acm.org/publications/policies/artifact-review-and-badging-current}, 2020.

¥bibitem{cobbe2021training}
Karl Cobbe, Vineet Kosaraju, Mohammad Bavarian, Mark Chen, Heewoo Jun,
Lukasz Kaiser, Matthias Plappert, Jerry Tworek, Jacob Hilton, Reiichiro Nakano,
Christopher Hesse, and John Schulman.
¥newblock Training verifiers to solve math word problems.
¥newblock ¥emph{arXiv:2110.14168}, 2021.

¥bibitem{guo2025deepseekr1}
Daya Guo et al.
¥newblock DeepSeek-R1: Incentivizing reasoning capability in LLMs via reinforcement learning.
¥newblock ¥emph{arXiv:2501.12948}, 2025.

¥bibitem{hu2024openrlhf}
Jian Hu, Xibin Wu, Weixun Wang, Xianyu, Dehao Zhang, and Yu Cao.
¥newblock OpenRLHF: An easy-to-use, scalable and high-performance RLHF framework.
¥newblock ¥emph{arXiv:2405.11143}, 2024.

¥bibitem{sheng2024verl}
Guangming Sheng et al.
¥newblock HybridFlow: A flexible and efficient RLHF framework.
¥newblock ¥emph{arXiv:2409.19256}, 2024.

¥bibitem{zhang2025verifybench}
Xiaoyu Zhang et al.
¥newblock VerifyBench: Benchmarking reference-based reward models for reinforcement learning from human feedback.
¥newblock In ¥emph{ICLR}, 2026.

¥end{thebibliography}

% =============================================================================
¥appendix
% =============================================================================

¥section{Compute Resources}
¥label{app:compute}

All experiment logs including per-step GPU utilization, memory consumption,
and wall-clock runtimes are available on Weights ¥& Biases at
¥url{https://wandb.ai/tinker-rl-lab/tinker-rl-lab-world-class}.
Model checkpoints are hosted on HuggingFace Hub at
¥url{https://huggingface.co/arvindcr4/tinker-rl-bench-*}.

¥begin{table}[h]
¥centering
¥caption{Compute allocation by platform and experiment group.}
¥begin{tabular}{llcc}
¥toprule
¥textbf{Platform} & ¥textbf{Experiment Group} & ¥textbf{¥# Experiments} & ¥textbf{Est. GPU-Hours} ¥¥
¥midrule
Tinker API & Scaling (Qwen3, Qwen3.5, Llama) & 5 & $¥sim$320 ¥¥
Tinker API & Frontier models (235B, DeepSeek, Nemotron) & 3 & $¥sim$280 ¥¥
Tinker API & Dense vs.‾MoE comparison & 2 & $¥sim$150 ¥¥
Tinker API & Cross-task tool-use & 2 & $¥sim$80 ¥¥
Tinker API & New architectures (GPT-OSS, Kimi-K2) & 2 & $¥sim$120 ¥¥
Modal A100 & PPO baselines (Qwen3-8B, Llama-8B) & 2 & $¥sim$96 ¥¥
Modal A100 & Full HumanEval evaluation & 1 & $¥sim$48 ¥¥
Modal A100 & KL divergence tracking & 1 & $¥sim$40 ¥¥
Modal A100 & Held-out evals (Qwen3-32B, Qwen3.5-27B) & 2 & $¥sim$66 ¥¥
¥midrule
 & ¥textbf{Total} & ¥textbf{20} & $¥sim$¥textbf{1,200} ¥¥
¥bottomrule
¥end{tabular}
¥end{table}

See ¥texttt{COMPUTE.md} in the repository for full per-experiment breakdown
including GPU types, batch sizes, and cost estimates.

¥section{Hyperparameter Tables}
¥label{app:hyperparams}

Full hyperparameter sweep ranges used in sensitivity analysis (Section‾¥ref{fig:sensitivity}):

¥begin{table}[h]
¥centering
¥caption{Hyperparameter sweep ranges for sensitivity analysis.}
¥begin{tabular}{lcc}
¥toprule
¥textbf{Parameter} & ¥textbf{Sweep Values} & ¥textbf{Default} ¥¥
¥midrule
Learning rate & $¥{10^{-5}, 5 ¥times 10^{-5}, 10^{-4}, 5 ¥times 10^{-4}, 10^{-3}¥}$ & $10^{-4}$ ¥¥
Clip range & $¥{0.05, 0.1, 0.2, 0.3, 0.5¥}$ & 0.2 ¥¥
Entropy coeff. & $¥{0.0, 0.001, 0.01, 0.05, 0.1¥}$ & 0.01 ¥¥
Discount $¥gamma$ & $¥{0.9, 0.95, 0.99, 0.995, 1.0¥}$ & 0.99 ¥¥
GAE $¥lambda$ & $¥{0.8, 0.9, 0.95, 0.98, 1.0¥}$ & 0.95 ¥¥
¥bottomrule
¥end{tabular}
¥end{table}

¥section{Additional Results}
¥label{app:results}

% Figure 9: Comparison bars
¥begin{figure}[h]
¥centering
¥includegraphics[width=0.85¥linewidth]{figures/comparison_bars.pdf}
¥caption{Final accuracy comparison across RL libraries on the arithmetic task,
sorted by performance. Error bars indicate $¥pm 1$ standard deviation across 5 seeds.
Dashed lines show floor (random baseline, 0.5¥%) and ceiling (oracle, 100¥%).}
¥label{fig:comparison_bars}
¥end{figure}

See ¥texttt{BASELINES.md} in the repository for complete floor/ceiling baseline documentation
and ¥texttt{BENCHMARKS¥_COMPARISON.md} for the full comparison table with existing RL benchmarks.

% =============================================================================
% NeurIPS Paper Checklist (Mandatory)
% =============================================================================
¥section*{NeurIPS Paper Checklist}

¥begin{enumerate}

¥item {¥bf Claims}
    ¥begin{enumerate}
        ¥item Do the main claims made in the abstract and introduction accurately reflect the paper's contributions and scope?
        ¥answerYes{See Section‾¥ref{sec:results} for experimental evidence supporting all three main claims.}
    ¥end{enumerate}

¥item {¥bf Limitations}
    ¥begin{enumerate}
        ¥item Does the paper discuss the limitations of the work performed by the authors?
        ¥answerYes{See Section‾¥ref{sec:limitations} discussing model scale, task coverage, platform dependency, LoRA-only evaluation, and statistical power limitations.}
    ¥end{enumerate}

¥item {¥bf Theory Assumptions and Proofs}
    ¥begin{enumerate}
        ¥item For each theoretical result, does the paper provide the full set of assumptions and a complete (or correct) proof?
        ¥answerNA{This is an empirical benchmarking paper; no theoretical results are claimed.}
    ¥end{enumerate}

¥item {¥bf Experimental Result Reproducibility}
    ¥begin{enumerate}
        ¥item Does the paper fully disclose all the information needed to reproduce the main experimental results of the paper to the extent that it affects the main claims and/or conclusions of the paper?
        ¥answerYes{Full code, Docker environments, seed management, and step-by-step commands in ¥texttt{REPRODUCE.md}. All experiments use 5 seeds with mean $¥pm$ SE and 95¥% CIs reported.}
    ¥end{enumerate}

¥item {¥bf Open access to data and code}
    ¥begin{enumerate}
        ¥item Does the paper provide open access to the data and code, with sufficient instructions to faithfully reproduce the main experimental results?
        ¥answerYes{Code: ¥url{https://github.com/arvindcr4/tinker-rl-lab}. Models: ¥url{https://huggingface.co/tinker-rl-bench}. All data, checkpoints, and scripts are publicly available under MIT license.}
    ¥end{enumerate}

¥item {¥bf Experimental Setting/Details}
    ¥begin{enumerate}
        ¥item Does the paper specify all the training and test details (e.g., data splits, hyperparameters, how they were chosen, and type of optimizer) necessary to understand the results?
        ¥answerYes{Section‾¥ref{sec:setup} and Table‾¥ref{tab:hyperparams} provide full hyperparameter settings. Appendix‾¥ref{app:hyperparams} details sweep ranges.}
    ¥end{enumerate}

¥item {¥bf Experiment Statistical Significance}
    ¥begin{enumerate}
        ¥item Does the paper report error bars suitably and correctly defined or other appropriate information about the statistical significance of the experiments?
        ¥answerYes{All results report mean $¥pm$ SE across 5 seeds with 95¥% confidence intervals. Performance profiles and bootstrap CIs via ¥texttt{rliable}.}
    ¥end{enumerate}

¥item {¥bf Experiments Compute Resources}
    ¥begin{enumerate}
        ¥item For each experiment, does the paper provide sufficient information on the computer resources (type of compute workers, memory, time of execution) needed to reproduce the experiments?
        ¥answerYes{Appendix‾¥ref{app:compute} describes GPU types (NVIDIA L4/T4), runtimes, and ¥texttt{COMPUTE.md} provides full breakdown.}
    ¥end{enumerate}

¥item {¥bf Code Of Ethics}
    ¥begin{enumerate}
        ¥item Does the research conducted in the paper conform, in every respect, with the NeurIPS Code of Ethics?
        ¥answerYes{No human subjects, no private data. Broader impact discussed in Section‾¥ref{sec:impact}.}
    ¥end{enumerate}

¥item {¥bf Broader Impacts}
    ¥begin{enumerate}
        ¥item Does the paper discuss both potential positive societal impacts and negative societal impacts of the work performed?
        ¥answerYes{Section‾¥ref{sec:impact} discusses how the benchmark promotes transparency and reproducibility while acknowledging compute access disparities.}
    ¥end{enumerate}

¥item {¥bf Safeguards}
    ¥begin{enumerate}
        ¥item Does the paper describe safeguards that have been put in place for responsible release of data or models that have a high risk for misuse?
        ¥answerNA{The benchmark evaluates existing public models on mathematical reasoning. No new capabilities with misuse potential are introduced.}
    ¥end{enumerate}

¥item {¥bf Licenses for existing assets}
    ¥begin{enumerate}
        ¥item Are the creators or original owners of assets (e.g., code, data, models), used in the paper, properly credited and are the license and terms of use explicitly mentioned and properly respected?
        ¥answerYes{All libraries cited with proper attribution. GSM8K‾¥cite{cobbe2021training} used under MIT license. Model licenses (Qwen: Apache 2.0, Llama: Community License) respected.}
    ¥end{enumerate}

¥item {¥bf New Assets}
    ¥begin{enumerate}
        ¥item Are new assets introduced in the paper well documented and is the documentation provided alongside the assets?
        ¥answerYes{Code released under MIT license with comprehensive README, REPRODUCE.md, ARTIFACT.md, and inline documentation.}
    ¥end{enumerate}

¥item {¥bf Crowdsourcing and Research with Human Subjects}
    ¥begin{enumerate}
        ¥item For crowdsourcing experiments and research with human subjects, does the paper include the full text of instructions given to participants and screenshots, if applicable, as well as details about compensation?
        ¥answerNA{No crowdsourcing or human subjects research.}
    ¥end{enumerate}

¥item {¥bf Institutional Review Board (IRB) Approvals or Equivalent for Research with Human Subjects}
    ¥begin{enumerate}
        ¥item Does the paper describe potential risks incurred by study participants, whether combensation was provided to study participants, and whether informed consent was obtained?
        ¥answerNA{No human subjects research.}
    ¥end{enumerate}

¥item {¥bf Declaration of LLM usage}
    ¥begin{enumerate}
        ¥item Does the paper describe the usage of LLMs if it is an important, original, or non-standard component of the core methods?
        ¥answerYes{LLMs (Qwen, Llama families) are the primary subject of evaluation. Their usage as RL training targets is described in Sections‾¥ref{sec:benchmark} and‾¥ref{sec:setup}. LLMs were not used for paper writing.}
    ¥end{enumerate}

¥end{enumerate}

¥end{document}
